---
title: "30 Years of Post-disturbance Recruitment in Tropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric MARCON
    authsuperscript: 1
  - name: Bruno HERAULT
    authsuperscript: 2
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB (Institut National Ploytechnique Félix Houphoüet Boigny) 
    address: Yamoussoukro, Ivory Coast
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >
  Résumé de l'article.
keywords: [Taxonomic and Functional Diversity, Recruitment, Resilience, Tropical Forests, Disturbance Dynamics]
journalinfo: Hal 00679993
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author 
output: 
  bookdown::gitbook: default
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::word_document2: default
---


```{r Options, echo=FALSE, message=FALSE,warning=FALSE}
knitr::opts_chunk$set(cache=T, warning=FALSE, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=40),fig.align="center")
# Installation des packages si nécessaire et chargement
Library <- function(Packages) {
  InstallAndLoad <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {install.packages(Package, repos="https://cran.rstudio.com/")}
    require(Package, character.only = TRUE)
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Ajouter les packages nécessaires ici
Library("kableExtra")
#source("Scripts/GithubPages.R")
source("Scripts/DrawingTools.R")
```

# Introduction

Determining the response of tropical forests to disturbance is a key to predict their fate in the global changing context.
In the last decades, tropical forests experienced a wide range of disturbance, from radical land-use changes for agriculture or mining [@Dezecache2017a;@Dezecache2017b] to more insidious changes of communities structure, diversity and functioning following anthropogenic activities like selective logging [@Baraloto2012a;@Herault2016] or climate change [@Aubry-Kientz2015]. 
In that respect a vast literature successfully modeled communities response to disturbance in terms of tree growth [@Gourlet-Fleury2000], tree height [@Rutishauser2016], carbon stocks and fluxes [@Putz2012;@Martin2015;@Piponiot2016]. 
Similar approaches regarding forest composition and diversity, however, have been hindered by the huge biological diversity, often focusing on common or mainly commercial species, and the scarcity of long-term monitoring [@Sebbenn2008;@Rozendaal2010;@Vinson2015].

Communites trajectories after disturbance, defined here as the evolution of communities diversity along time, depend on the trees surviving from before disturbance and on the new trees recruited afterward [@Herault2018]. 
Surviving trees proved to mirror the composition of pre-disturbance forest so communities' diversity response is driven by the diversity and composition of recruited trees, which build the future community and determine the resilience of pre-disturbance state.
Recruitment trajectories depend first on the composition and diversity of initial community that partly condition the available pool of recruited species [@Herault2018].
Then trajectories depend on recruitment processes either stochastic, like random dispersal, recruitment and death [@Hubbell2001], or deterministic like niche-based competition processes [@Adler2007] 
While stochastic processes would build communities similar to random samples of larger regional metacommunities, deterministic processes rely on the abiotic environment and filter-out recruited species according to their ecology. 
Understand communities response to disturbance comes back first to estimate the balance between these determinants, the initial composition and the different recruitment processes, and then to elucidate communities resilience and the maintenance of the initial differences among local communities [@Diaz2005;@Gardner2007;@Schwartz2017].

The processes shaping communities response will differently affect the functional diversity [@Kunstler2016], that considers species functioning and ecology [@Violle2007b], and the taxonomic diversity, which consider all species equal.
The coupling between functional and taxonomic trajectories will then be insightful for the identification of ecological processes at stake, and communities resilience should be addressed considering both functional and taxonomics characteristics [@Fukami2005]. 
Functional trajectories of recruited trees are shaped by the deterministic processes that are oriented towards the use of limiting resources. 
Recruitment processes will then either filter-out the les efficient functional strategies and restrict the functional diversity, or exclude the less competitive ones and increase the functional diversity though the competition among species and limitation of their functional similarity [@Ackerly2003;@McGill2006]. 
The disturbance response of tropical forests where the light is limiting would then be a shift from slow-growing, long-living species with "conservative" resource use to fast growing, resource "acquisitve" species [@Denslow1980;@Molino2001;@Bongers2009]. 
To detect these processes the large functional trait-based litterature identified key leaf, wood and life-history functional traits assessing species ecology and resources acquisition strategy [@Wright2004;@Chave2009b;@Herault2011].

In this paper we follow the fate of a recruited tree communities (60121 individuals) over 30 years on a large disturbance gradient, with 10 to 60% of forest biomass removed. 
We assess the taxonomic and functional diversity of recruited trees, using a large functional trait database covering the leaf, wood and life-history spectra. 
We compare the observed trajectories to null models representing random trees recruitment and randomized functional traits. 
We aimed (i) to assess the role of deterministic processes compared to stochastic recruitment after disturbance, (ii) assess the taxonomic and functional convergence of forest communities and the maintenance of taxonomic composition in the long term, and (iii) determine the degree of resilience of the ecosystem.

# Material and Methods

## Study Site

The Paracou station is located in a lowland tropical rain forest in French Guiana (5°18'N and 52°53'W). 
Climate is tropical wet with mean annual precipitation averaging 2980 mm.y^-1^ (30-y period) and a 3-months dry season (< 100 mm.months^-1^) from mid-August to mid-November, and a one-month dry season in March [@Wagner2011]. 
Elevation ranges from 5 to 50 m and mean annual temperature is 26°C. 
Soils are thin acrisols over a layer of transformed saprolite with low permeability generating lateral drainage during heavy rains. 
The disturbance experiment spread over a network of twelve 6.25ha plots (Table \@ref(tab:Tab1)) that underwent three disturbance treatments in 1986-1987 [@Herault2018]. 
Dominant families are Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae. 

```{r Tab1, echo=FALSE,as.is=T}
Intervention <- data.frame(
  Treatment=c("Control", "T1", "T2", "T3"),
  Timber=c("", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha",
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha"),
  Thinning=c("", 
             "", 
             "DBH $\\geq$ 40 cm, non-valuable species, $\\approx$ 30 trees/ha", 
             "DBH $\\geq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha  "),
  Fuelwood=c("", 
             "", 
             "", 
             "40 cm $\\leq$ DBH $\\leq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha"),
  AGBLost=c(0, "$[12\\%-33\\%]$", "$[33\\%-56\\%]$", "$[35\\%-56\\%]$")
)
names(Intervention)[5] <- "\\%AGB lost"
knitr::kable(Intervention, caption="Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.", longtable = FALSE, booktabs = TRUE, escape = FALSE,format="latex") %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = TRUE)
```

## Inventories Protocol and Dataset Collection 

All trees above 10 cm DBH were mapped and measured annually since 1984. 
During inventories, trees were first identified with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. 
Botanical campaigns have been carried out every 5 to 6 years from 2003 onwards. 
These changes in identification protocol raised methodological issues as vernacular names usually correspond to different botanical species, resulting in significant taxonomic uncertainties that were propagated to composition and diversity metrics. 
Vernacular names were replaced through multinomial trials $M_v\Big(\big[s_1, s_2, …, s_N\big],\big[\alpha_1, \alpha_2,…, \alpha_3\big]\Big)$  based on the observed association probability $\big[\alpha_1, \alpha_2,…, \alpha_3\big]$ between each vernacular name *v* and the species $\big[s_1, s_2, …, s_N\big]$ recorded in the inventory. 
See appendix 1 and @Aubry-Kientz2013 for the detailed methodology. 
To avoid remaining identification caveats, the simulated botanical inventories were reported at genus level.

Eight functional traits were considered, representing leaf economics (leaves thickness, toughness, total chlorophyll content and specific leaf area), wood economics (wood specific gravity and bark thickness) and life history traits (maximum specific height and seed mass).
Traits were exctracted from the BRIDGE project [^1] where trait values were measured on nine forest plots infrench guianan, including two in Paracou. 
Missing trait values of the trait database (10%) were filled by multivariate imputation by chained equation using the Mice r package [@Mice2011].
As traits variability was lower within genus and families, we accounted for the phylogenetic signal of the functional traits by restricting the gap filling processes to samples pertaining to the next higher taxonomic level. 
As seed mass information corresponded to a classification into discrete mass classes, no data filling process was applied so analysis were performed only considering the 414 botanical species of the seed mass dataset.

[^1]: http://www.ecofog.gf/Bridge/

## Recruitment trajectories

To disentangle the recruitment processes from overall dynamics, communities were split into per-disturbance surviving trees and those recruited since disturbance.
Recruited communities were examined either considering the "punctual recruitment", *i.e.* recruited trees by 2-year intervals, or all recruits since disturbance as the “accumulated recruits”. 
Eventually, in disturbed plots the recruited communities were examined distinguishing the undisturbed and logging gap areas to test the validity of recruitment processes for the whole area.

The taxonomic diversity was assessed through Richness and the Hill number translation of Shannon and Simpson indices [@Hill1973;@chao2015estimating;@Marcon2015b].  
The three diversities belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (*q*), which grasps the balance between richness and evenness in the community through the value of *q* that emphasizes common species.
Functional trajectories were estimated with the Rao quadratic entropy and completed by the trajectories of traits community weighted means (CWM), representing the average trait value in a community weighted by relative abundance of the species carrying each value [@Diaz2007;@Garnier2004].
Seed mass trajectories were reported by the proportion of each class recorded in the inventories.
The similarity between the recruited trees and the pre-disturbance forest was measured with the turnover metrics detailed in @Podani2013a.
To estimate the importance of stochastic processes the recruitment was compared to the trajectories of a random sampling.
For the taxonomic trajectories the random sampling was a shuffle of trees among plots that preserved species abundance and tree density, and for the functional diversity it was a shuffling of functional trait values among species.

All composition and diversity metrics correspond to the median and 90% percentile obtained after 50 iterations of the taxonomy uncertainty propagation framework and the gap filling process.
The stochastic trajectories were similarily obtained after 50 iterations of the random sampling.

# Results

## Recruitment Diversity

### Taxonomic Diversity

The diversity trajecctories of punctual recruitment followed a consistent trajectory after disturbance with first an increase of the richness and a decrease of the evenness (Figure (\@ref(fig:DivTraj)). 
For all disturbed plots,both richness and evenness tended to return towards initial values but none had recovered 30 years after disturbance.
The accumulated recruits displayed sharp increasing richness (order 0) and decreasing evenness (order 2) after intense disturbance (T3 and some T2, Appendix I, fig. S1).

```{r DivTraj, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Trajectories of Richness, Shannon and Simpson diversity for 2-years laps punctual  recruitment (upper panels) and divergence to null model (lower panels). Lines colors refer to the perturbation regime: green for control, blue for T1, orange for T2 and red for T3 disturbance treatments. Plain lines correspond to the median observed after uncertainty propagation and are given along with the 95\\% confidence interval (grey envelope).", fig.env="figure*"}
load("DB/RecruitmentPunctual");load("DB/RecruitmentPunctual_Nullmodel_Diff")

TrajectoryDiffNull(RecDB=RecPun,RecDB_Diff=RecPun_DiffNull)

```

Punctual and accumulated recruitment diversities were then compared to the stochastic trajectories of a random sampling. 
Richness (order 0) and evenness (order 2) of punctual recruits remained equivalent or higher than for a random sampling in control plots while both were lower in disturbed plots.
Disturbed plots however followed humped-shaped trajectories heading towards a recovery of the initial state (Figure \@ref(fig:DivTraj)). 
Accumulated recruitment richness and evenness were higher or equivalent to those of the random sampling after low disturbance intensity (plots T1 and some plots T2) but lower after intense disturbance (plots T3 and a plot T2, Appendix I fig. S1).

### Functional Diversity and Composition

Communities functional diversity was measured with the Rao diversity and compared to the stochastic trajectories of a random traits shuffling. 
In disturbed plots (T2 and T3), the functional diversity decreased until 15 years after disturbance (Figure \@ref(fig:FunTraj)) before recovering towards the initial values. 
While the recovery was not achieved for the most disturbed plots, the functional diversity of lighter disturbance plots recovered faster and for some T1 plots exceeded the initial values.
For all plots, disturbed or not, the observed functional diversity was lower than this of the random model, to the exception of two plots T1.

```{r FunTraj, fig.asp=1.1/1, echo=FALSE,fig.cap="Functional diversity of punctual recruited trees from the considered functional traits and divergence to null model. Values reported correspond to the plot-level median and the 95\\% confidence interval obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes and 50 run of the null model. Lines colors correspond to the logging treatment initially applied (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/RecruitmentPunctual_Functional");load("DB/RecruitmentPunctual_Functional_Nullmodel_Diff")

TrajectoryRec_fun(RecDB_fun=RecPun_Fun,RecDB_fun_Diff=RecPun_Fun_DiffNull)
```

Trajectories of the functional traits showed a switch in disturbed plots towards species with large exchange surface area, light tissues (high SLA, low leaf toughness and thickness and low wood specific gravity) with smaller maximum height (Figure \@ref(fig:CWM)). 
Functional traits either followed hump-shaped trajectories with an ongoing recovery or an achieved return to the initial state (for SLA,Bark thickness and leaf thickness and Hmax to a certain extent).

```{r CWM, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Community weighted means (CWM) of the four disturbance treatment for the four leaf traits, the two stem traits  and the specific Hmax. Values reported correspond to the plot-level median obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes. Lines colors correspond to the disturbance intensity (green for control, blue for T1,orange for T2 and red for T3).", fig.env="figure*"}
load("DB/Recruits_CWM")

par(mfrow=c(2,4),mar=c(3,3,2,1),oma=c(2,1,4,1),no.readonly=TRUE)
PlotCWM(Rec_CWM)
legendCWM()
mtext("Punctual recruits CWM",line=2,adj=0,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.6,cex=0.8,line=-0.2,outer=TRUE)
```


## Recruitment Turnover

In control plots species turnover remained highly stable for the 30 sampled years (Figure \@ref(fig:Turnover)), reflecting a strong similarity between the initial plots composition and the punctual recruits. 
In disturbed plots, the taxonomic turnover followed a marked hump-backed trajectory, with a maximum value reached around 15 years after disturbance and a maximum positively correlated to the disturbance intensity ($\rho_{spearman}=0.93$). 
Thirty years after disturbance the turnover of all disturbed plots had return to low values close to zero.

```{r Turnover, echo=FALSE,fig.cap="Trajectories over the 30 sampled years of the abundance-based turnover between recruited trees and intial communities before disturbance. Grey envelopes correspond to the 0.025 and 0.975 percentiles of the uncertainty propagation procedue and lines to the median in green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/Turnover_toInit")

turnover(Turn)
par(mar=c(1,2,1,0))
mtext("Recruitment specific turnover compared to initial",side=3,adj=0,cex=1.2)
mtext("Years since disturbance",side=1,adj=1,line=-2)
mtext("Equivalent diversity",side=2,line=1)
```


# Discussion

## On the underlyings of the hump-shaped trajectories

The trajectories of recruitment richness, key functional traits (SLA and bark thickness) and the species turnover exhibited clear hump-shaped, unimodal trajectories. 
Within communities disturbed and undisturbed areas followed simultaneous trajectory, supporting an impact of disturbance at the whole plot scale.

Trees recruited in the first place (0-8 years) resembled the pre-disturbance communities and their functional diversity matched this of a stochastic recruitment process.
This first recruitment phase likely involved already grown saplings (DBH <10cm) that germinated before disturbance and immediatly benefitted from the increased enlightment and the alleviated competition following diturbance [@Herault2010]. 

Then during a second phase the evenness and the functional diversity of recruited trees decreased, translating a more restricted pool of recruited species based on their functional strategy.
Following intense disturbance, sharp changes in the SLA, wood density and leaf thickness trajectories then revealed the prominent recruitment of short-lived, fast growing hard pionneers species with competitive and efficient light acquisition (Figure @ref(fig:Turnover)) [@Wright2004;@Chave2009b;@Herault2011;@Reich2014].  
The second phase therefore likely incorporated true recruits, *i.e.* trees germinated from the seed bank, submitted to deterministic recruitment processes excluding the least competitive species.
Along time diversity trajectories returned towards the initial state and matched the values of stochastic recruitment, reflecting a decreasing importance of the deterministic recruitment processes.
The balance between deterministic and stochastic processes was determined by the initial disturbance intensity.
After light disturbance (T1 plots), the recruitment matched the composition of the initial communities but the pool of recruited species was restricted by the competitive exclusion for resources. 
dominant species were more pioneers and light demanders, with strategies of efficient resource acquisition (high SLA and leaf chlorophyll content) and inexpensive, short-lived tissues (low leaf thickness and thoughness, small Hmax and low wood density and bark thickness).
Although competitive exclusion decreased the functional diversity of recruited trees, their eveness remained high, so the were no high dominance of hard pionners [@Hubbell1999;@Sheil2003;@Bongers2009].
This might due to recruitment and dispersal limitation due to the short dispersal distance observed for tropical trees, specifically in Paracou with the genetic clumping of some pioneers [@Leclerc2015;@Scotti2015a]. 
After intense disturbance however, the recruitment composition rapidly differed from the pre-disturbance state and corresponded to a sharp increase of the SLA and bark thickness CWM, likely reflecting the overhelming recruitment of hard pioneers. 
This recruitment phase lasted for 15 years after disturbance, which corresponds to the life expectancy of hard pionners, and was followed by a progressive increase of the functional diversity. 
The recruitment remained dominated by acquisitive functional strategies but the initial composition and diversity progressively recovered. Communities trajectories involved an interplay between stochastic and deterministic recruitment, advocating that disturbance effectively maintained forests diversity [@Molino2001;@Sheil2003].


## On the resilience of the recruitment process

After 30 years, although taxonomic and functional diversity had recovered initial values, the recruitment remained more restricted than with the stochastic recruitment of undisturbed plots. 
Despite the settlement after intense disturbance of long-lived pioneers, recruitment processes proved resilient but slow to recover. 

The recovery of recruitment processes meant the convergence of communities towards the initial state.
More than commonly thought, the taxonomic trajectory of ecosystems would then rely on their pre-disturbance characteristics and concur to maintain the initial taxonomic differences among local communities [@Anderson2007]. 
In contrast the trajectories of traits and functional diversity were essentially similar among treatments, arguing for the confluence of communities in the functional space despite their divergence in taxonomic composition [@Fukami2005].
This confirmed previous results from the Paracou experiment, conducted 10 years [@Molino2001] and 20 years [@Baraloto2012a] after disturbance, where the early signs of the resilience of taxonomic and functional composition had been detected. 
Still, the functional diversity and several traits CWM values remained altered 30 years after disturbance, suggesting the long-term impact of disturbance. 
Both taxonomic and functional recovery therefore spread over long periods, i.e. several decades, and to involve the germination of trees from the seed bank and therefore affect the stock of recruitable species and communities response and resilience to additional disturbance [@Norden2009]. 

# Conclusion

The hindsight of the 30 years of forest monitoring highlighted a two-phased disturbance response, distinguished by the balance between stochastic and determinisitic recruitment processes.
Communities trajectories were first driven by the stochastic recruitment of already-grown saplings mirroring the predisturbance state before it is dominated by true recruits from the seed bank selected based on their resource acquisition strategy.
Both communities taxonomy and functioning recovered in the long term but while all communities followed similar trajectories in the funcional space, their taxonomic trajectories relied upon communities composition and diversity and maintained initial local differences.
Although resilient within the range of applied disturbance, communities followed a decade-long recovery and probably underwent compositional changes of their seed bank which entailed great caution regarding the forest conservation and exploitation guidelines if the pursued objectives are a complete recovery of pre-disturbance ecosystem properties.

----

`r if (knitr:::is_html_output()) '# References {-}'`
