---
title: "30 Years of Recruitment in Tropical Forest After Selective Logging"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric MARCON
    authsuperscript: 1
  - name: Bruno HERAULT
    authsuperscript: 2
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB (Institut National Ploytechnique Félix Houphoüet Boigny) 
    address: Yamoussoukro, Ivory Coast
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >
  Résumé de l'article.
keywords: [mot-clés, séparés par des virgules]
journalinfo: Hal 00679993
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
bibliography: references.bib
lang: english
keywordlabel: Mots-clés # Keywords
corrauthorlabel: Contact # Corresponding author
output:
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::gitbook: default
  bookdown::word_document2: default
---

<!-- Options de knitr et de R (librairies, ...) --> 

```{r Options, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(cache=F, warning=FALSE, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=40),fig.align="center")
# Installation des packages si nécessaire et chargement
Library <- function(Packages) {
  InstallAndLoad <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {install.packages(Package, repos="https://cran.rstudio.com/")}
    require(Package, character.only = TRUE)
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Ajouter les packages nécessaires ici
Library("kableExtra")
#source("Scripts/GithubPages.R")
source("Scripts/DrawingTools.R")
```

# Introduction

Determining the response of tropical forests to disturbance is a key to predict their fate in a global change context.
A vast literature has successfully modeled the response of tropical forest dynamics, carbon stocks and fluxes to anthropogenic and natural disturbances [@Gourlet-Fleury2000;@Putz2012;@Martin2015;@Piponiot2016]. However, regarding diversity, similar attempts have been hindered by both the huge biological diversity and the scarcity of long-term monitoring. Although the community composition response to disturbance was identified for common species assemblages, this approach is usually confined to few commercial and valuable species [@Sebbenn2008;@Rozendaal2010;@Vinson2015]. Given the complexity of adressing simultaneously the response of dozens of species, community-scale approaches have been developped [@deAvila2016].

The post-disturbance community dynamic is largely determined by the recruitment process, i.e. from seed production, dispersion and germination to seedlings' and saplings' growth until a threshold, usually 10cm DBH in tropical forest plots.
Major ecological processes, such as environmental filtering or limiting similarity, occur before formal recruitment and determine the taxonomic composition of trees eventually integrating the forest community [@Ackerly2003]. The recruitment process is inherently linked to the disturbance regime that locally changes ecosystem's biotic and abiotic conditions and creates opportunities for pioneer species to strive in the long term rather than undergoing the competitive exclusion for resources in mature patches [@Denslow1980]. This relies on the Intermediate Disturbance Hypothesis (IDH) that explains the maintenance of tropical forests biodiversity by the patchy variability of environmental conditions in space and time [@Guitet2018]. This disturbance-induced environmental variability, when not too intense, is assumed to enlarge the ecological range of species present in the community [@Molino2001;@Bongers2009] and shapes its taxonomic diversity, vegetative structure, physiology as well as carbon, nutrients, and water cycles [@Anderson-Teixeira2013]. Empirical tests of the IDH in tropical rainforests, though, proved hard to succeed and yielded controversial results [@Hubbell1999;@Molino2001;@Sheil2003].

During post-disturbance times, the shift from resource-acquisitive to resource-conservative ecological strategies may be detected in leaves (leaf thickness, toughness, chlorophyll content and specific area) and stem (wood specific gravity and bark thickness) and life-history traits (maximum height at adult stage and class of seed mass) [@Wright2004;@Chave2009b;@Herault2011]. The relative importance of recruitment of new individuals and of mortality of distrubance survivors will shape the new forest and its functioning. Given that disturbance survivors largely mirror the pre-disturbance forest composition [@Piponiot2018], predicting the recruitment comosition and diversity trajectory would be a major step towards the prediction of the future of tropical forest in a changing global environment where dusturbance are expected to become more and more frequent. This would give insights into the resilience of this hyperdiverse ecosystems, elucidate the determinism, or not, of tropical forests trajectories, test the convergence after disturbance of taxonomic and functional communities towards initial state and also help future adaptative conservation strategies [@Diaz2005;@Gardner2007;@Schwartz2017].

In this paper we follow the fate of a recruited tree communities (XX individuals) over 30 years on a disturbance gradient, with 10 to 60% of forest biomass removed. We assessed the taxonomic as well as functional diversity of recruited trees, using a large functional trait database covering, the leaf, wood and life-history spectra. We aimed to (i) assess the role of environmental filtering selecting the recruited trees according to their competitivity for resource acquisition, (ii) resolve the convergence of communities and the maintenance of taxonomic composition in the long term, and (iii) determine the global resilience of the ecosystem.

# Material and Methods


## Study Site

The Paracou station is located in a lowland tropical rain forest in French Guiana (5°18'N and 52°53'W). Climate is tropical wet with mean annual precipitation averaging 2980 mm.y^-1^ (30-y period) and a 3-months dry season (< 100 mm.months^-1^) from mid-August to mid-November, and a one-month dry season in March [@Wagner2011]. Elevation ranges from 5 to 50 m and mean annual temperature is 26°C. Soils are thin acrisols over a layer of transformed saprolite with low permeability generating lateral drainage during heavy rains. The disturbance experiment spraid over a network of twelve 6.25ha plots (Table \@ref(tab:Tab1)) that underwent three disturbance treatments in 1986-1987 [@Herault2018]. Dominant families are Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae. 

```{r Tab1, echo=FALSE}
Intervention <- data.frame(
  Treatment=c("Control", "T1", "T2", "T3"),
  Timber=c("", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha",
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha"),
  Thinning=c("", 
             "", 
             "DBH $\\geq$ 40 cm, non-valuable species, $\\approx$ 30 trees/ha", 
             "DBH $\\geq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha  "),
  Fuelwood=c("", 
             "", 
             "", 
             "40 cm $\\leq$ DBH $\\leq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha"),
  AGBLost=c(0, "$[12\\%-33\\%]$", "$[33\\%-56\\%]$", "$[35\\%-56\\%]$")
)
names(Intervention)[5] <- "\\%AGB lost"
knitr::kable(Intervention, caption="Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.", longtable = FALSE, booktabs = TRUE, escape = FALSE) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = TRUE)
```

## Inventories Protocol and Dataset Collection 
All trees above 10 cm DBH were mapped and measured annually since 1984. During inventories, trees were first identified with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. Botanical campaigns have been carried out every 5 to 6 years from 2003 onwards. This raised methodological issues as vernacular names usually correspond to different botanical species, resulting in significant taxonomic uncertainties that were propagated to composition and diversity metrics. Vernacular names were replaced through multinomial trials $M_v\Big(\big[s_1, s_2, …, s_N\big],\big[\alpha_1, \alpha_2,…, \alpha_3\big]\Big)$  based on the observed association probability $\big[\alpha_1, \alpha_2,…, \alpha_3\big]$ between each vernacular name *v* and the species $\big[s_1, s_2, …, s_N\big]$ recorded in the inventory. See appendix 1 and @Aubry-Kientz2013 for the detailed methodology. To avoid remaining identification caveats, the simulated botanical inventories were reported at genus level.

Six functional traits, representing leaf economics (leaves thickness, toughness, total chlorophyll content and specific leaf area, the leaf area per unit dry mass), wood economics (wood specific gravity and bark thickness) and life history traits (maximum specific height and seed mass), come from the BRIDGE project [^1] where trait values were measured on nine french guianan forest plots, including two in Paracou. Missing trait values (X%) were filled using multivariate imputation by chained equation (mice) restricted to samples pertaining to the next higher taxonomic level, in order to account for the phylogenetic signal of the functional traits (refs MICE). As seed mass information corresponds to a classification into mass classes, no data filling process was applied so analysis were performed only considering the 414 botanical species of the seed mass dataset.

[^1]: http://www.ecofog.gf/Bridge/

Functional trajectories were estimated with the Rao quadratic entropy using community weighted means (CWM) [@Diaz2007;@Garnier2004]. Seed mass trajectories were reported by the proportion of each class recorded in the inventories.
All composition and diversity metrics are the average obtained after 50 iterations of uncertainty propagation (taxonomy and trait values). 


## Recruitment trajectories

We split the forest community in 'survivors, i.e. trees that survived the disturbance,  and post-disturbance recruited trees. Two recruitment metrics were examined: on the one hand the "punctual recruitment" by 3-year intervals after disturbance, on the other hand all recruited trees since disturbance, hereafter “accumulated recruits”. The taxonomic diversity was assessed through bias corrected Richness and the Hill number translation of Shannon and Simpson indices [@Hill1973;@chao2015estimating;@Marcon2015b]. These three indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (q), which grasps the balance between richness and evenness in the community with common species weighting more than rare ones when q increases. The similarity between the recruited trees and the pre-disturbance forest was measured with the turnover metrics detailed in @Podani2013a. To determine whether recruitment trajectories ensued from a pure random process the observed diversity trajectories were compared to those generated by 50 repetitions of a null model that shuffled individuals among plots while preserving species abundance and plots’ tree density. 

# Results

## Recruitment Diversity

### Punctual and accumulated recruits’ diversity 

The diversity of trees recruited in T1, T2 and T3 disturbed plots followed an asymptotique trajectory, first displaying higher richness and lower evenness than those observed in control plots and then lower richness and lower evenness.

> The OLS analysis proved a significant time-dependence of the punctual recruitment diversity for the three diversity order, Richness, Shannon and Simpson and a significant effect of the logging treatment (P<0.01 in any case).
> It besides proved a significant time-dependence of the accumulated recruitment diversity for the three order and a significant effect of the logging.


```{r Fig1, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Trajectories for the 30 sampled years of Richness, Shannon and Simpson diversity for 2-years laps punctual recruitment (upper graphics) and the divergence of observed trajectories to the null model (lower graphics). Lines colors correspond to the perturbation regime: green for control, blue for T1, orange for T2 and red for T3 disturbance treatments. Lines correspond to the median observed after 50 iteration of the the taxonomic uncertainty propagation for observed values and 50 run of the null model run, both trajectories are given along with the 95\\% confidence interval (grey envelope).", fig.env="figure*"}
load("DB/RecruitmentPunctual");load("DB/RecruitmentPunctual_Nullmodel_Diff")

TrajectoryDiffNull(RecDB=RecPun,RecDB_Diff=RecPun_DiffNull)

```

```{r Fig2, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Trajectories for the 30 sampled years of Richness, Shannon and Simpson diversity of communities recruited since disturbance (upper graphics) and the divergence of observed trajectories to the null model (lower graphics). Lines colors correspond to the perturbation regime: green for control, blue for T1, orange for T2 and red for T3 disturbance treatments. Lines correspond to the median observed after 50 iteration of the the taxonomic uncertainty propagation for observed values and 50 run of the null model run, both trajectories are given along with the 95\\% confidence interval (grey envelope).", fig.env="figure*"}
load("DB/RecruitmentAccum");load("DB/RecruitmentAccum_Nullmodel_Diff")

TrajectoryDiffNull(RecDB=RecAccum,RecDB_Diff=RecAccum_DiffNull)

```

### Comparison to null model

Punctual and accumulated recruitment diversity of orders 0, 1 and 2 were compared to a null recruitment model corresponding to the random sampling of recruited trees holding constant the number of trees recruited.
For control plots the richness and evenness of observed punctual recruitment were constantly equivalent or higher than obtained with the null model (Figure \@ref(fig:Fig2)).
For the disturbed plots the richness of observed punctual recruitment remained equivalent to that of the null model but the evenness that was first equivalent or higher became and remained lower than that of the null model from 10 years after disturbance whenever the disturbance intensity.

In the same way as for punctual recruitment the richness and evenness of accumulated recruits in control plots was equivalent or higher than observed for the null model for all the post logging survey. 
On disturbed plots the richness of accumulated recruits followed that of the null model whereas the evenness, initially higher,  became lower than that of the null model after some time (Figure \@ref(fig:Fig3)). 


### Recruitment functional diversity

The dataset available encompassed 88% of all species recorded in the recruitment inventories.
Because the species for which trait values were unavailable were not evenly distributed among plots and along time so we decided to discard them from the inventories to preserve homogeneous data filling.

Punctual and accumulated recruitment diversity of order 0, 1 and 2 were compared to a null recruitment model corresponding to the random shuffle of complete set of traits value among species (Figure \@ref(fig:Fig4)).


```{r Fig4, echo=FALSE,fig.cap="Functional diversity of punctual recruited trees from the considered functional traits. Values reported correspond to the plot-level median and the 95\\% confidence interval obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes. Lines colors correspond to the logging treatment initially applied (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/RecruitmentPunctual_Functional")

TrajectoryRec_fun(RecDB_fun=RecPun_Fun,RecDB_fun_Diff=RecPun_Fun_DiffNull)
```

Recruits functional diversity displayed a unimodal response to disturbance with a maximum positively correlated to disturbance intensity but with a time at maximum that was unrelated to the disturbance treatment.

### Recruitment traits weighted means

The trajectories of recruited trees functional and life traits weighted means showed the dominance after disturbance of species displaying large exchange surface area and light tissues (high SLA, low leaf toughness and low wood specific gravity) (Figure \@ref(fig:Fig5)).
all traits trajectories dosplayed asymptotic CWM trajectories, except the SLA that displayed a unimodal trajectory with a maximum positively correlated to disturbance intensity and a time a maximum around ten years after disturbance equivalent for all treatments. 

```{r Fig5, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Community weighted means (CWM) of the four disturbance treatment for the four leaf traits, the two stem traits  and the specific Hmax. Values reported correspond to the plot-level median obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes. Lines colors correspond to the disturbance intensity (green for control, blue for T1,orange for T2 and red for T3).", fig.env="figure*"}
load("DB/Recruits_CWM")

par(mfrow=c(2,4),mar=c(3,3,2,1),oma=c(2,1,4,1),no.readonly=TRUE)
PlotCWM(Rec_CWM)
legendCWM()
mtext("Punctual recruits CWM",line=2,adj=0,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.6,cex=0.8,line=-0.2,outer=TRUE)
```


## Recruitment Turnover

The abundance turnover of recruited trees compared to the initial stand inventoried before exploitation remained even for the 30 sampled years in the control plots (Figure \@ref(fig:Fig6)). 
For the disturbed plot the abundance turnover displayed a unimodal response to disturbance, with maximum reached around 10 years after disturbance that was positively correlated to the disturbance intensity. 
The turnover has returned close to zero for all plots 30 years after disturbance.


```{r Fig6, out.width = '50%',echo=FALSE,fig.cap="Trajectories over the 30 sampled years of the abundance-based turnover between recruited trees and intial stands before disturbance. 50 repetitions of the taxonomic propagation process returned the distribution of trunover values, grey envelopes correspond to the 0.025 and 0.975 percentiles and lines to the median in green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/Turnover_toInit")

turnover(Turn)
par(mar=c(1,2,1,0))
mtext("Recruitment specific turnover compared to initial",side=3,adj=0,cex=1.2)
mtext("Years since disturbance",side=1,adj=1,line=-2)
mtext("Equivalent diversity",side=2,line=1)
```


# Discussion

From the 30 years of forest dynamics survey in the Paracou station we highlighted contrasting recruitment mechanisms between control and post-disturbance tropical forests. 
The difference among treatments were also significant which suggest an intensity threshold between the first and second disturbance treatment and could involve gaps size, the richness of exploited trees or the intervention density [@Denslow1980].

> If we go for a model, maybe we can test the most relevant parameter!?

Disturbance increased the recruitment rate which subsequently increased the richness of recruitment but impaired its evenness, all the more so that disturbance intensity was high. 
After disturbance the recruitment was dominated by a restricted pool of species and Shannon and Simpson diversities decreased down to lower values than those of control plots.

For the 15 first years following disturbance the recruitement seemed driven by the growth of pre-disturbance saplings on hold of suitable growth condition that benefitted from the environmental changes. 
This translated by an increase in the functional diversity of recruits for all disturbance intensities and an even distribution of functional traits.
According to the high increase of SLA CWM trajectory (Figure \@ref(fig:Fig5)) right after disturbance these early recruited species were pionneers dominating the recruited community until 10 years after disturbance.

After this first recruitment phase, in the case of low disturbance intensity (T1 plots) the recruitment was highly nested within the community in place and displayed a richness and evenness similar to those of a random sampling null-model. 
This argues for the existence of some recruitment limitation meaning the strive of a variety of species, even when not the most fitted for the environmental context, because the best competitors couldn't set in the disturbed area.
The key role of recruitment limitation in recovery dynamic was already highlighted for low disturbance intensity in tropical wet forests [@Hubbell1999;@Sheil2003;@Bongers2009] and supported the IDH as it mitigates competitive exclusion and maintains some inferior competitiors in the community.
Recruitment and dispersal limitation are consistent with the short dispersal distance observed for tropical trees, and it was already highlighted in Paracou through the genetic clumping observed for some pioneers [@Leclerc2015;@Scotti2015a].

In the case of high disturbance intensity, the species turnover was high and the functional diversity decreased rapidly, along a recruitment shift towards acquisitive functional strategies. 
This was particularly marked for functional traits of the leaf and stem and for life history traits (Leaf thickness, SLA, Wood Specific Gravity and Hmax) which values represented a community of fast-growing species with a high resources acquisition strategies [@Wright2004;@Chave2009b;@Herault2011;@Reich2014].
The time elpased between disturbance and the significance of those functional and diversity shifts suggests that the second recruitment phase stems from the germination at the moment of the disturbance of seeds of the seed bank. 
This recruitment would undergo a range of environmental filters and thus restricting the pool of species recruited [@Molino2001]. 

Eventually, we observed that disturbance induced particular recruitment mechanisms but also set aside mechanisms at stake in natural plots.
Diversity of recruited community in undisturbed plots was higher than this observed for a random-sampling null model, suggesting mechanisms maintaining high diversity like negative density dependance [@Harms2000].


Previous studies already highlighted the significant diversity increase of communities as a whole (Mirabel et al., in.prep.) that seemed supported here by a shift in recruitment dynamics in favor of acquisitive functional types [@TerSteege2001].

After 30 years stands functional characteristics had recovered whatever the disturbance intensity but significant taxonomic diversity difference persisted when the disturbance intensity was high. 
Stands functioning would then recover faster than taxonomic composition, which questions the recovery and the role of functional redundancy regarding tropical forests diversity.
In any case, the recruitment turnover compared to initial stand ended up close to zero.
This argued for the maintenance or for the recovery in relation to the stand [@Anderson2007] of the before-disturbance environment.
It also supported the convergence of communities tree species taxonomic and functional composition in the long term [@Li2016].
The communities recruited after disturbance would then persist for more than 30 years as suggested by the remaining difference among treatments but the initial recruitment dynamics seemed restored at this point, ensuring communities resilience if no other disturbance is applied.
The different results of two studies on the same site 10 years [@Molino2001] and 20 years [@Baraloto2012a] after logging already suggested the recovery towards initial taxonomic and functional condition of the stands in revealing a decreasing proportion of pioneers between the two periods.
This is however only valid for a single disturbance event, given that second recruitment phase assumed to rely on the seed bank triggers a storage effect likely to modify the trajectories after another disturbance event.
We assume that in this case the competitive exclusion among dormant life-stage (seeds or even seedlings) would be harsher and bring more radical changes in the recruitment composition and functional profile.
Some functional shift towards more acquisitive, small seeded and globally disturbance-resitant species building a more resilient community as it has been observed by @Haddad2008 who linked the persistence of species to their ability to recover from disturbance which is the seed mass, growth rate linked to WD and SLA for plants.
Additionally to a somehow depleted seed bank this strongly questioned the repeatability of observed trajectories after other disturbance events.

Our long-term study of diversity trajectories after disturbance disentangled the mechanisms underlying stands diversity. 
While, in undisturbed forests, mechanisms like negative density dependence enhanced species diversity, any disturbance induced new mechanisms depending on the disturbance intensity.
In any case, communities short term response first reflected the growth of a variety of already grown saplings which increased the taxonomic and functional evenness of the community.
When the disturbance was low, recruitment limitation was balanced against environmental filters and induced a recruited community mirroring the stand in place while selecting more acquisitive functional types.
When the disturbance was intense however the species turnover was high and induced a recruited community significantly dominated by acquisitive functional types and a restricted pool of species.
In both cases plots recruitment is specific, because it is determined by the surrounding stand in case of low disturbance or because the competitive exclusion randomly selected a restricted pool of species after intense disturbance.
At the landscape scale, then, disturbance increased the diversity through the "between patch" difference [@Denslow1987;@Chesson2000; @Sheil2003;@Guitet2017].

> Connell 1978, refute the "equal chance HP because trees don't recruit equally"
> Contrary to Hubbell and Molino: the question is "do gaps allow specific species to establish which would thrive otherwise?" [@Sheil2003]. The turnover answers that yes it does
<!-- Baraloto, 2012: on juveniles the differences were still at stake. --> 

----

`r if (knitr:::is_html_output()) '# References {-}'`
