---
title: "30 Years of Post-disturbance Recruitment in Tropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric MARCON
    authsuperscript: 1
  - name: Bruno HERAULT
    authsuperscript: 2
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB (Institut National Ploytechnique Félix Houphoüet Boigny) 
    address: Yamoussoukro, Ivory Coast
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >
  Résumé de l'article.
keywords: [Taxonomic and Functional Diversity, Recruitment, Resilience, Tropical Forests, Disturbance Dynamics]
journalinfo: Hal 00679993
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author 
output: 
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::gitbook: default
  bookdown::word_document2: default
---


```{r Options, echo=FALSE, message=FALSE,warning=FALSE}
knitr::opts_chunk$set(cache=T, warning=FALSE, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=40),fig.align="center")
# Installation des packages si nécessaire et chargement
Library <- function(Packages) {
  InstallAndLoad <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {install.packages(Package, repos="https://cran.rstudio.com/")}
    require(Package, character.only = TRUE)
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Ajouter les packages nécessaires ici
Library("kableExtra")
#source("Scripts/GithubPages.R")
source("Scripts/DrawingTools.R")
```

# Introduction

Determining the response of tropical forests to disturbance is a key to predict their fate in the global changing context.
In the last decades, tropical forests experienced a wide range of disturbance, from radical land-use changes for agriculture or mining [@Dezecache2017a;@Dezecache2017b] to more insidious changes of communities structure, diversity and functioning following anthropogenic activities like selective logging [@Baraloto2012a;@Herault2016] or climate change [@Aubry-Kientz2015]. 
In that respect a vast literature successfully modeled communities response to disturbance in terms of tree growth [@Gourlet-Fleury2000], tree height [@Rutishauser2016], carbon stocks and fluxes [@Putz2012;@Martin2015;@Piponiot2016]. 
Similar approaches regarding forest composition and diversity, however, have been hindered by the huge biological diversity, often focusing on common or mainly commercial species, and the scarcity of long-term monitoring [@Sebbenn2008;@Rozendaal2010;@Vinson2015].

Communites trajectories after disturbance, defined here as the evolution of communities diversity along time, depend on the trees surviving from before disturbance and on the new trees recruited afterward [@Herault2018]. 
Surviving trees proved to mirror the composition of pre-disturbance forest so communities' diversity response is driven by the diversity and composition of recruited trees, which build the future community and determine the resilience of pre-disturbance state.
Recruitment trajectories depend first on the composition and diversity of initial community that partly condition the available pool of recruited species [@Herault2018].
Then trajectories depend on recruitment processes either stochastic, like random dispersal, recruitment and death [@Hubbell2001], or deterministic like niche-based competition processes [@Adler2007] 
While stochastic processes would build communities similar to random samples of larger regional metacommunities, deterministic processes rely on the abiotic environment and filter-out recruited species according to their ecology. 
Understand communities response to disturbance comes back first to estimate the balance between these determinants, the initial composition and the different recruitment processes, and then to elucidate communities resilience and the maintenance of the initial differences among local communities [@Diaz2005;@Gardner2007;@Schwartz2017].

The processes shaping communities response will differently affect the functional diversity [@Kunstler2016], that considers species functioning and ecology [@Violle2007], and the taxonomic diversity, which consider all species equal.
The coupling between functional and taxonomic trajectories will then be insightful for the identification of ecological processes at stake, and communities resilience should be addressed considering both functional and taxonomics characteristics [@Fukami2005]. 
Functional trajectories of recruited trees are shaped by the deterministic processes that are oriented towards the use of limiting resources. 
Recruitment processes will then either filter-out the les efficient functional strategies and restrict the functional diversity, or exclude the less competitive ones and increase the functional diversity though the competition among species and limitation of their functional similarity [@Ackerly2003;@McGill2006]. 
The disturbance response of tropical forests where the light is limiting would then be a shift from slow-growing, long-living species with "conservative" resource use to fast growing, resource "acquisitve" species [@Denslow1980;@Molino2001;@Bongers2009]. 
To detect these processes the large functional trait-based litterature identified key leaf, wood and life-history functional traits assessing species ecology and resources acquisition strategy [@Wright2004;@Chave2009b;@Herault2011].

In this paper we follow the fate of a recruited tree communities (60121 individuals) over 30 years on a large disturbance gradient, with 10 to 60% of forest biomass removed. 
We assess the taxonomic and functional diversity of recruited trees, using a large functional trait database covering the leaf, wood and life-history spectra. 
We compare the observed trajectories to null models representing random trees recruitment and randomized functional traits. 
We aimed (i) to assess the role of deterministic processes compared to stochastic recruitment after disturbance, (ii) assess the taxonomic and functional convergence of forest communities and the maintenance of taxonomic composition in the long term, and (iii) determine the degree of resilience of the ecosystem.

# Material and Methods

## Study Site

The Paracou station is located in a lowland tropical rain forest in French Guiana (5°18'N and 52°53'W). 
Climate is tropical wet with mean annual precipitation averaging 2980 mm.y^-1^ (30-y period) and a 3-months dry season (< 100 mm.months^-1^) from mid-August to mid-November, and a one-month dry season in March [@Wagner2011]. 
Elevation ranges from 5 to 50 m and mean annual temperature is 26°C. 
Soils are thin acrisols over a layer of transformed saprolite with low permeability generating lateral drainage during heavy rains. 
The disturbance experiment spread over a network of twelve 6.25ha plots (Table \@ref(tab:Tab1)) that underwent three disturbance treatments in 1986-1987 [@Herault2018]. 
Dominant families are Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae. 

```{r Tab1, echo=FALSE,as.is=T}
Intervention <- data.frame(
  Treatment=c("Control", "T1", "T2", "T3"),
  Timber=c("", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha",
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha"),
  Thinning=c("", 
             "", 
             "DBH $\\geq$ 40 cm, non-valuable species, $\\approx$ 30 trees/ha", 
             "DBH $\\geq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha  "),
  Fuelwood=c("", 
             "", 
             "", 
             "40 cm $\\leq$ DBH $\\leq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha"),
  AGBLost=c(0, "$[12\\%-33\\%]$", "$[33\\%-56\\%]$", "$[35\\%-56\\%]$")
)
names(Intervention)[5] <- "\\%AGB lost"
knitr::kable(Intervention, caption="Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.", longtable = FALSE, booktabs = TRUE, escape = FALSE,format="latex") %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = TRUE)
```

## Inventories Protocol and Dataset Collection 

All trees above 10 cm DBH were mapped and measured annually since 1984. 
During inventories, trees were first identified with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. 
Botanical campaigns have been carried out every 5 to 6 years from 2003 onwards. 
These changes in identification protocol raised methodological issues as vernacular names usually correspond to different botanical species, resulting in significant taxonomic uncertainties that were propagated to composition and diversity metrics. 
Vernacular names were replaced through multinomial trials $M_v\Big(\big[s_1, s_2, …, s_N\big],\big[\alpha_1, \alpha_2,…, \alpha_3\big]\Big)$  based on the observed association probability $\big[\alpha_1, \alpha_2,…, \alpha_3\big]$ between each vernacular name *v* and the species $\big[s_1, s_2, …, s_N\big]$ recorded in the inventory. 
See appendix 1 and @Aubry-Kientz2013 for the detailed methodology. 
To avoid remaining identification caveats, the simulated botanical inventories were reported at genus level.

Eight functional traits were considered, representing leaf economics (leaves thickness, toughness, total chlorophyll content and specific leaf area), wood economics (wood specific gravity and bark thickness) and life history traits (maximum specific height and seed mass).
Traits were exctracted from the BRIDGE project [^1] where trait values were measured on nine forest plots infrench guianan, including two in Paracou. 
Missing trait values of the trait database (10%) were filled by multivariate imputation by chained equation using the Mice r package [@Mice2011].
As traits variability was lower within genus and families, we accounted for the phylogenetic signal of the functional traits by restricting the gap filling processes to samples pertaining to the next higher taxonomic level. 
As seed mass information corresponded to a classification into discrete mass classes, no data filling process was applied so analysis were performed only considering the 414 botanical species of the seed mass dataset.

[^1]: http://www.ecofog.gf/Bridge/

## Recruitment trajectories

To disentangle the recruitment processes from overall dynamics, communities were split into per-disturbance surviving trees and those recruited since disturbance.
Recruited communities were examined either considering the "punctual recruitment", *i.e.* recruited trees by 2-year intervals, or all recruits since disturbance as the “accumulated recruits”. 
Eventually, in disturbed plots the recruited communities were examined distinguishing the undisturbed and logging gap areas to test the validity of recruitment processes for the whole area.

The taxonomic diversity was assessed through Richness and the Hill number translation of Shannon and Simpson indices [@Hill1973;@chao2015estimating;@Marcon2015b].  
The three diversities belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (*q*), which grasps the balance between richness and evenness in the community through the value of *q* that emphasizes common species.
Functional trajectories were estimated with the Rao quadratic entropy and completed by the trajectories of traits community weighted means (CWM), representing the average trait value in a community weighted by relative abundance of the species carrying each value [@Diaz2007;@Garnier2004].
Seed mass trajectories were reported by the proportion of each class recorded in the inventories.
The similarity between the recruited trees and the pre-disturbance forest was measured with the turnover metrics detailed in @Podani2013a.
To estimate the importance of stochastic processes the recruitment was compared to the trajectories of a random sampling.
For the taxonomic trajectories the random sampling was a shuffle of trees among plots that preserved species abundance and tree density, and for the functional diversity it was a shuffling of functional trait values among species.

All composition and diversity metrics correspond to the median and 90% percentile obtained after 50 iterations of the taxonomy uncertainty propagation framework and the gap filling process.
The stochastic trajectories were similarily obtained after 50 iterations of the random sampling.

# Results

## Recruitment Diversity

### Taxonomic Diversity

Punctual recruits' diversity followed a consistent trajectory among disturbance treatments with first higher richness and equivalent evenness than in control plots and then equivalent richness and lower evenness (Figure (\@ref(fig:Fig1)). 
For accumulated recruits, the richness (order 0) in highly disturbed plots (T3 and some T2) was higher than in control plots, consistently with the higher richness of punctual recruits, and the evenness (order 2) was lower, specifically for the most disturbed plots (Appendix I, fig. S1).

```{r Fig1, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Trajectories of Richness, Shannon and Simpson diversity for 2-years laps punctual  recruitment (upper panels) and divergence to null model (lower panels). Lines colors refer to the perturbation regime: green for control, blue for T1, orange for T2 and red for T3 disturbance treatments. Plain lines correspond to the median observed after uncertainty propagation and are given along with the 95\\% confidence interval (grey envelope).", fig.env="figure*"}
load("DB/RecruitmentPunctual");load("DB/RecruitmentPunctual_Nullmodel_Diff")

TrajectoryDiffNull(RecDB=RecPun,RecDB_Diff=RecPun_DiffNull)

```

Punctual and accumulated recruitment diversity of orders 0, 1 and 2 were then compared to a null random recruitment model. In control plots the richness (order 0) and evenness (order 2) of punctual recruits remained equivalent or higher than for the null random model. In contrast, for all disturbance levels both richness and evenness were lower than these of a random null model but displayed a significant but unachieved humped-shaped trajectory in all plots (Figure \@ref(fig:Fig1)). Accumulated recruitment richness and evenness were higher or equivalent to those of the null model for plots T1 and some plots T2 but lower for plots T3 and a plot T2 (AppendixI, fig. S1).

### Functional Diversity and Composition

The functional diversity (Rao diversity) of punctual recruitment was measured and compared to a null model of random traits shuffling. 
In the most disturbed plots (T2 and T3), the functional diversity was decreasing and lower to control plots until 15 years after disturbance (Figure \@ref(fig:Fig3)). 
It then increased to values equivalent or higher to those observed in control plots. 
For all disturbed and control plots the observed functional diversity trajectories over 30 years end lower than for the null model of random traits shuffle, except for two T1 plots.


```{r Fig2, fig.asp=1.1/1, echo=FALSE,fig.cap="Functional diversity of punctual recruited trees from the considered functional traits and divergence to null model. Values reported correspond to the plot-level median and the 95\\% confidence interval obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes and 50 run of the null model. Lines colors correspond to the logging treatment initially applied (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/RecruitmentPunctual_Functional");load("DB/RecruitmentPunctual_Functional_Nullmodel_Diff")

TrajectoryRec_fun(RecDB_fun=RecPun_Fun,RecDB_fun_Diff=RecPun_Fun_DiffNull)
```

Trajectories of recruited trees in the functional spaces showed the dominance after disturbance of species displaying large exchange surface area and light tissues (high SLA, low leaf toughness and thickness and low wood specific gravity) (Figure \@ref(fig:Fig3)). All traits trajectories displayed univariate CWM trajectories with leaf toughness, wood specific gravity and bark thickness, decreasing before stabilizing at low values around 15 after disturbance, except SLA and leaf thickness that, for most disturbed plots, displayed a unimodal trajectory with a maximum reach around 15 years after disturbance. 

```{r Fig3, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Community weighted means (CWM) of the four disturbance treatment for the four leaf traits, the two stem traits  and the specific Hmax. Values reported correspond to the plot-level median obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes. Lines colors correspond to the disturbance intensity (green for control, blue for T1,orange for T2 and red for T3).", fig.env="figure*"}
load("DB/Recruits_CWM")

par(mfrow=c(2,4),mar=c(3,3,2,1),oma=c(2,1,4,1),no.readonly=TRUE)
PlotCWM(Rec_CWM)
legendCWM()
mtext("Punctual recruits CWM",line=2,adj=0,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.6,cex=0.8,line=-0.2,outer=TRUE)
```


## Recruitment Turnover

In control plots species turnover remained highly stable for the 30 sampled years (Figure \@ref(fig:Fig4)), reflecting a strong similarity between the initial plots composition and the punctual recruits. 
In disturbed plots, turnover displayed a unimodal response to disturbance, with a maximum value reached around 15 years and with a value positively correlated to the disturbance intensity ($\rho_{spearman}=0.93$). 
The turnover trajectory returned close to zero for all plots 30 years after disturbance.

```{r Fig4, echo=FALSE,fig.cap="Trajectories over the 30 sampled years of the abundance-based turnover between recruited trees and intial communities before disturbance. Grey envelopes correspond to the 0.025 and 0.975 percentiles of the uncertainty propagation procedue and lines to the median in green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/Turnover_toInit")

turnover(Turn)
par(mar=c(1,2,1,0))
mtext("Recruitment specific turnover compared to initial",side=3,adj=0,cex=1.2)
mtext("Years since disturbance",side=1,adj=1,line=-2)
mtext("Equivalent diversity",side=2,line=1)
```


# Discussion

The composition, diversity, and functional trajectories in post-disturbance times was analysed ine the Paracou forest ecosystems to determine their resilience and help future management strategies. The 30 years-long monitoring trajectories allowed to identify the different steps of forests recovery and the corresponding assembly rules. We identified three distinct phases. The first corresponds to the recruitment of pre-disturbance saplings that reflected the diversity and composition of the pre-disturbance community composition. The second was driven by "true recruits" germinating from the seed bank in the early post-disturbance times. The third involved environmental selection and progressive competition among species according to the disturbance intensity, which build recruited communities more or less divergent from the initial stand and persisting in the medium term.

## On the underlyings of the hump-shaped trajectories

The trajectories of punctual recruitment richness, of some key functional traits (SLA and bark thickness) and of species turn-over exhibited clear hump-shaped, unimodal trajectories. 

Indeed, trees recruited immediately after disturbance did not substantially differ from the pre-disturbance communities because they come from the growth of already-established saplings 
(DBH <10cm) that benefit from the increased light availability and alleviated competition [@Herault2010]. 
In this phase (0-8 years), the functional diversity of the recruitment remained stable, equating these of control plots and matching stochastic recruitment processes. 
Then, depending on disturbance intensity, the recruitment incorporated true recruits, i.e. trees germinated from the seed bank, resulting in a decrease of recruited trees evenness and 
functional diversity, i.e. a restriction in the pool of recruited species. 
Following intense disturbance, sharp changes in the SLA, wood density and leaf thickness trajectories can be observed and this reveals the prominent recruitment, 
above an intensity threshold and after a couple of years, of short-lived, fast growing hard pionneers species with competitive and efficient light acquisition (Figure @ref(fig:Fig4)) [@Wright2004;@Chave2009b;@Herault2011;@Reich2014].  

Following this first phase, the recruitment progressively incorporated true recruits, *i.e.* trees germinated from the seed bank, resulting in a decrease of recruited trees evenness and 
functional diversity. 
XXXX not clear in the following Then an interplay between deterministic processes excluding less competitive and stress-tolerant species, 
and stochastic recruitment as in mature forests that progressively emerged again. 
The balance betwwen both processes was determined by the intensity of the initial disturbance.XXXXX 
After low disturbance intensity (T1 plots), the taxonomic composition of the recruitment resembled the pre-disturbance communities but pool of recruited species was restricted by 
selective pressures on the strategy of resource use. 
Favored species were more pioneers and light demanders, with strategies of efficient resource acquisition (high SLA and leaf chlorophyll content) and inexpensive, short-lived tissues 
(low leaf thickness and thoughness, small Hmax and low wood density and bark thickness). 
Correspondingly the functional diversity increased, equating or exceeding this of control plots, which revealed an overdispersion of the functional traits that was probably driven by 
the limitation of functional similarity among species. 
Despite some competitive exclusion then, recruited species still browsed a large functional range with no high dominance of hard pionners [@Hubbell1999;@Sheil2003;@Bongers2009].
This might due to recruitment and dispersal limitation due to the short dispersal distance observed for tropical trees, specifically in Paracou with the genetic clumping of some pioneers 
[@Leclerc2015;@Scotti2015a]. 
After intense disturbance however, the first species recruited were hard pioneers, which taxonomic composition highly differed from the initial community and that generated a sharp 
increase for the SLA and bark thickness weighted values. 
This was observed for 15 years after disturbance, which corresponds to the life expectancy of hard pionners, before the functional diversity increased again, acquisitive functional 
still dominating, and the recruited trees progressively resembled the initial composition and diversity. 
Communities trajectories then involved the interplay between stochastic recruitment and deterministic processes and advocated the role of disturbance to maintain forests diversity, 
in line with the Intermediate Disturbance Hypothesis [@Molino2001;@Sheil2003].


## On the resilience of the recruitment process

Thirty years, both taxonomic and functional diversity had recovered pre-disturbance levels, reaching equivalent levels as those of undisturbed forests. However, the recruitment remain more important in disturbed plots and did not match the richness and evenness of stochastic recruitment even if the divergence from stochasticity progressively shrinked, arguing for a consistent but long-term recovery of the recruitment process. Eventually, the recruitment should thus converge towards the initial state, with the recruited species durably mirroring the pre-disturbance community. More than commonly thought, the ecosystem trajectories in the taxonomic space depends on the pre-disturbance ecosystem characteristics and this concur to maintain the initial taxonomic differences among local communities [@Anderson200]. In contrast the trajectories of traits and functional diversity were essentially similar among treatments, arguing for the confluence of communities in the functional space despite their divergence in taxonomic composition [@Fukami2005].

The taxonomic and functional characteristics of tropical forest ecosystems were thus both resilient, even if the settlment of long-lived pioneers after intense disturbance still impact communities functioning and diversity 30 years after disturbance.
This confirms previous results from the Paracou experiment, conducted 10 years [@Molino2001] and 20 years [@Baraloto2012a] after disturbance, where the ea  rly signs of the resilience of taxonomic and functional composition had been detected. The time length, i.e. several decades, of the recovery processes entails great caution regarding the forest conservation and exploitation guidelines if the pursued objectives are a complete recovery of pre-disturbance ecosystem properties.

????To that was added the involvement of the seed bank in the recovery processes which own resilience remains unknown.

????Any storage effect, altering the pool of recruitable species, would modify the resilience of the community itself and impact the response to further disturbance events [@Norden2009]. 

????In such case, the competitive exclusion among dormant life-stage (seeds or even seedlings) would be harsher and likely bring more radical changes in the recruitment composition and functional profile of the community.

# Conclusion

The 30 years monitoring of the Paracou plots highlighted the tropical forests' response to disturbance composed of two recruitment phase modulated by the disturbance intensity. In the short-term forests response was driven by the enhanced growth of grown saplings benefiting from the alleviated competition and the environmental changes.  Above an intensity threshold the recruitment was besides dominated by hard-pioneers radically changing the recruitment composition, diversity and, likely, functioning. In the long-term response was driven by recruits from the seed bank which underwent selection towards light demanding species and similarity limitation enhancing the functional diversity. These deterministic processes followed a gradual balance with the stochastic recruitment of mature forests which eventually restored communities diversity and composition, maintaining their initial differences. Although forests proved resilient to intense disturbance this appeared to be a long-term processes likely only valid for single disturbance events.

----

`r if (knitr:::is_html_output()) '# References {-}'`
