---
title: 30 Years of Recruitment in Tropical Forest After Selective Logging 
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric MARCON
    authsuperscript: 1
  - name: Bruno HERAULT
    authsuperscript: 2
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB (Institut National Ploytechnique Félix Houphoüet Boigny) 
    address: Yamoussoukro, Ivory Coast
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >
  Résumé de l'article.
keywords: [mot-clés, séparés par des virgules]
journalinfo: Hal 00679993
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{amsmath}
bibliography: references.bib
lang: french # english
keywordlabel: Mots-clés # Keywords
corrauthorlabel: Contact # Corresponding author
output:
  bookdown::gitbook
#   bookdown::pdf_book:
#     base_format: EcoFoG::article
#     keep_tex: TRUE
#  bookdown::word_document2
---

<!-- Options de knitr et de R (librairies, ...) --> 

```{r Options, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(cache=F, warning=FALSE, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=40),fig.align="center")
# Installation des packages si nécessaire et chargement
Library <- function(Packages) {
  InstallAndLoad <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {install.packages(Package)}
    require(Package, character.only = TRUE)
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Ajouter les packages nécessaires ici
#source("Scripts/GithubPages.R")
source("Scripts/DrawingTools.R")
```

<!-- Début du document ci-dessous --> 


# Introduction

Recruitment, the demographic process whereby trees germinate, grow and survive until adult stage, is a major determinent of forest dynamics after disturbance. 
Disturbance by definition changes ecosystems biotic and abiotic context and creates a temporal variability allowing a variety of species to strive despite the exclusion for resource competition [@denslow1980].
This mechanism is the driver of the Intermediate Disturbance Hypothesis (IDH) which explains the structure and the high biodiversity of tropical forests by the patchy variability of environmental conditions. 
This, to the extent that changes are not too drastic and excludes a portion of the ecological range, favors the growth and survival of numerous species browsing a wide ecological range [@Molino2001;@Bongers2009].<this extent remains to be found>.
These dynamics determine the vegetative structure of the forests, the ecosystem physiology, the carbon and nutrients cycling and the maintenance of species diversity [@Anderson-Teixeira2013].
This makes recruitment dynamics the drivers of major dynamics, such as the tree fall gaps dynamics which continuously impulses different recruitment regime between enlightened, dry gaps and shaded, wet undercover areas. 

Forests population dynamics were the focus of a large literature which successfully modeled the response of key forest features and functions such as carbon sequestration, commercial stocks, and diameter distribution dynamics in the short and in the long term [@Gourlet-Fleury2005].
However, the response of communities on their whole and the underlying assembly rules remains unknown. 
Insights lready highlighted a combination of environmental filters and stochastic events as drivers of the response to disturbance, but the importance of recruitment dynamics have scarcely been disentangled. This hampers to have a refined understanding of response dynamics and to accurately forecast of future forests dynamics.
If the regeneration of some species is successfully modeled, the response of the whole recruited community after disturbance but this approach requires an accurate knowledge of species ecology and behavior.
It would not be workable for the community on its whole,given the number of species in tropical forests and the variability of growth and survival in the early life stage among species.

In this paper we aimed to have some insights into the role of demographic dynamics underlying the response of tropical forest after disturbance. 
From the 30 years survey of a tropical forest after disturbance following a gradient of disturbance intensity, we aimed to (i) identify the evolution of the taxonomic and functional diversity and the functional characteristics of commnities after disturbance, (ii) highlight the environmental filters ruling the recruitment according to the disturbance intensity, and (iii) highlight the recovery of communities and the existence of cyclic dynamics for the taxonomic and functional diversity.

-	Hypotheses: 

  1) The more intense the logging, the most significant the impact on composition and diversity metrics. 

  2) We expect an increase in the stand diversity (Intermediate disturbance hypothesis) or no change in richness, at least for low logging intensity

  3) Exploitation would change the functional composition toward more light-demanding strategies 

  4) We expect the diversity to follow a cyclic dynamic, as observed for stands biomass, and the ecosystem functioning to be resilient, at least for low intensity treatment


# Material and Methods


*Study Site* 


Analyses were based on the inventories conducted at the Paracou station in French Guiana (5°18'N and 52°53'W) in a lowland tropical rain forest. 
The site corresponds to a tropical wet climate with mean annual precipitation averaging 2980 mm.y-1 (30-y period) with a 3-months dry season (< 100 mm.months-1) from mid-August to mid-November, and a one-month dry season in march. 
Elevation ranges between 5 and 50 m and mean annual temperature is 26°C. 
Soils correspond to thin acrisols over a layer of transformed saprolite with low permeability generating lateral drainage during heavy rains.
The experiment corresponds to a network of twelve 6.25ha plots where three logging treatments representing a gradient of logging intensity were applied from 1985 to 1987. 
Logging treatments were attributed according to a randomized plot design with three replicate blocks of four plots. 
The logging treatments correspond to averages of 10 trees remove per hectare with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees above 40 cm DBH for treatment 3 (T3). 
Treatments T2 and T3 included besides logging of 10trees/ha the thinning by poison girdling.


*Inventories Protocol and Dataset Collection* 


The study site corresponds to a tropical rain forest with a dominance of Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae. 
In the twelve experimental plots all trees above 10 cm DBH are mapped and measured annually since 1984, which allows to survey the trees recruited annually. 
During the inventories trees are first identified with a vernacular name, assigned by the field team, and then with a scientific name, assigned by a botanist during regular botanical campaigns. 
In 1984, specific vernacular names were given to 62 commercial or common species whereas two identifiers separating palm trees and trees were given to all other species. 
Botanical campaigns to identify all trees at the species level started in 2003 and continued every 5 to 6. 
In total from 1984 to 2016 around 850 species were identified in the 12 plots. 
For the trees that were identified at the species level vernacular names attributed in the first place generate some taxonomic uncertainty as they may correspond to a variety of botanical species. 
To account for this taxonomic uncertainty in the composition and diversity metrics we used a Bayesian framework estimating the expectancy and variance of composition and diversity metrics. 
The frameworks is based on a set of complete inventories extrapolated from the initial one through the replacement of vernacular name by botanical species. 
These replacement were made through a multinomial distribution M~v~([s~1~, s~2~, …, s~N~] ,[α~1~, α~2~,…, α~N~]) based on the observed association probability [α~1~, α~2~,…, α~N~] between each vernacular name v and the species recorded in the inventory [s~1~, s~2~, …, s~N~]. 
See appendix 1 for the detailed methodology. 
Diversity metrics correspond to the expectancy obtained after 50 iteration of the taxonomic uncertainty propagation framework. 


*Recruitment Survey* 


To assess the demographic dynamics along the 30 sampled year we divided the inventoried trees between trees recorded before the logging in 1984 and the trees recruited afterward. 
Analysis first surveyed the communities of trees recruited every 3 years-laps after exploitation, hereafter called “punctual recruitment”. 
Then to assess the diversity of the stands under settlement analysis surveyed the communities of recruited trees accumulated since exploitation, hereafter called “accumulated recruits”.
The taxonomic diversity was assessed through Richness and the Hill number translation of Shannon and Simpson indices [@Hill1973]. To tackle the unequal number of recruited trees among treatments the indices bias corrected estimator were used, following [@chao2015estimating;@Marcon2015b].
These three indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (q), which proved well suited for diversity studies [@Patil1982;@Tothmeresz1995]. 
This order q grasps the balance between richness and evenness in the community as it determines the emphasis on common species in the diversity metric, with common species weighting more than rare ones when q increases. 
The similarity between the recruited trees and the old growth forest was measured with the turnover metrics detailed in @Podani2013a. 
The metric used correspond to the relativized abundance replacement, the sum of abundance in one site that is replaced by completely different species, normalized by the maximum abundance shared by the two communities.

$$ T_{ab}=\frac{\displaystyle\sum_{i=1}^{n}|x_i^a - x_i^b| - \bigg|\sum_{i=1}^{n} x_i^a - \sum_{i=1}^{n} x_i^b \bigg|}{\displaystyle\sum_{i=1}^{n}max\big({x_i^a;x_i^b\big) }} $$

To determine whether trees recruitment ensued from a random process the observed diversity trajectories were compared to those generated by 50 repetitions of stochastic null models.
The null model for taxonomic diversity randomly shuffled individuals among plots while preserving species abundance and plots’ tree density. 


# Results

Over time, 104423 trees and 697 botanical species were recorded. 

**Recruitment Diversity** 


*Punctual and accumulated recruits’ diversity* 


The diversity of trees recruited in T1, T2 and T3 logged plots first constituted a community with higher richness and lower evenness than in control plots until 25 years after disturbance. From then on, disturbed plots displayed lower richness and lower evenness than control ones. 
*The OLS analysis proved a significant time-dependence of the punctual recruitment diversity for the three diversity order, Richness, Shannon and Simpson and a significant effect of the logging treatment (P<0.01 in any case). 
It besides proved a significant time-dependence of the accumulated recruitment diversity for the three order and a significant effect of the logging.*


```{r Fig1, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Trajectories for the 30 sampled years of Richness, Shannon and Simpson diversity for **(a)**  3-years laps punctual recruitment and **(b)** recruits accumulated since disturbance. Solid lines (green for control, blue for T1, orange for T2 and red for T3 disturbance treatments) correspond to the median observed after 50 iteration of the taxonomic uncertainty propagation, along with the 95% confidence interval (grey envelope)."}
load("DB/RecruitmentAccum");load("DB/RecruitmentPunctual")

par(mfrow=c(2,3),mar=c(2,3,1,1),oma=c(2,3,1,1),no.readonly=TRUE)
TrajectoryRec(RecPun,0.8)
TrajectoryRec(RecAccum,1)
mtext("(a) Richness",at=0.1,line=-0.6,outer=TRUE);mtext("(b) Shannon",at=0.44,line=-0.6,outer=TRUE)
mtext("(c) Simpson",at=0.76,line=-0.6,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.85,cex=0.8,line=1,outer=TRUE)
mtext("Equivalent diversity",side=2,padj=1,cex=0.8,line=0.8,outer=TRUE)
mtext("Punctual",side=2,adj=0.8,line=1.75,outer=TRUE)
mtext("Accumulated",side=2,adj=0.2,line=1.75,outer=TRUE)

```


*Comparison to null model* 


Punctual and accumulated recruitment diversity of order 0, 1 and 2 were compared to a null recruitment model corresponding to the random sampling of recruited trees holding constant the effectives of trees recruited.
For control plots the richness and evenness of observed punctual recruitment were constantly equivalent or higher than obtained with the null model \@ref(fig:Fig2). For the disturbed plots the richness of observed punctual recruitment remained equivalent to this of the null model, however the evenness that was first equivalent or higher became lower than this of the null model from 10 years after disturbance whenever the disturbance intensity.

```{r Fig2, out.width = '80%', fig.asp = 1.5/2, echo=FALSE, fig.cap="Punctual recruitment diversity trajectories for the Richness, Shannon and Simpson diversity compared to the null model trajectories. Values reported correspond to the median diversity observed after 50 repetition of the taxonomic uncertainty propagation. Solid lines correspond to the observed trajectories and dotted lines to the null model simulations.Line colors stand for the disturbance treatment with green for control, blue for T1, orange for T2 and red for T3 disturbance treatments.Values reported correspond to the median and 95% confidence intervalobserved after 50 iteration of the taxonomic uncertainty propagation and null model simulation."}
load("DB/RecruitmentPunctual_Nullmodel")

TrajectoryRec_vsNull(RecPun,RecPun_Null)
mtext("Punctual recruitment",line=1.2,adj=0,outer=TRUE,cex=1.2)
mtext("Richness",at=0.1,line=-0.5,outer=TRUE);mtext("Shannon",at=0.44,line=-0.5,outer=TRUE)
mtext("Simpson",at=0.76,line=-0.5,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.85,cex=0.8,line=1,outer=TRUE)
mtext("Equivalent diversity",side=2,padj=1,cex=0.8,line=1.5,outer=TRUE)
```


In the same way as for punctual recruitment the richness and evenness of accumulated recuits in control plots was equivalent or higher than observed for the null model for all the post logging survey. On disturbed plaots the richness of accumulated recruits followed this of the null model whereas the evenness, initially higher,  became lower than this of the null model after some time \@ref(fig:Fig3). 

```{r Fig3, out.width = '80%', fig.asp = 1.5/2, echo=FALSE, fig.cap="Accumulated recruitment diversity trajectories for the Richness, Shannon and Simpson diversity compared to the null model trajectories. Values reported correspond to the median diversity observed after 50 repetition of the taxonomic uncertainty propagation. Solid lines correspond to the observed trajectories and dotted lines to the null model simulations.Line colors stand for the disturbance treatment with green for control, blue for T1, orange for T2 and red for T3 disturbance treatments.Values reported correspond to the median and 95% confidence intervalobserved after 50 iteration of the taxonomic uncertainty propagation and null model simulation.."}
load("DB/RecruitmentAccum_Nullmodel")

TrajectoryRec_vsNull(RecAccum,RecAccum_Null)
mtext("Accumulated recruitment",line=1.2,adj=0,outer=TRUE,cex=1.2)
mtext("Richness",at=0.1,line=-0.5,outer=TRUE);mtext("Shannon",at=0.44,line=-0.5,outer=TRUE)
mtext("Simpson",at=0.76,line=-0.5,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.85,cex=0.8,line=1,outer=TRUE)
mtext("Equivalent diversity",side=2,padj=1,cex=0.8,line=1.5,outer=TRUE)
```


*Recruitment functional diversity * 


Punctual and accumulated recruitment diversity of order 0, 1 and 2 were compared to a null recruitment model corresponding to the random shuffle of complete set of traits valueamong species.


```{r Fig4, out.width = '50%', echo=FALSE,fig.cap="Functional diversity of punctual recruited trees from the considered functional traits. Values reported correspond to the plot-level median and the 95% confidence interval obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes. Lines colors correspond to the logging treatment initially applied (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/RecruitmentPunctual_Functional")

TrajectoryRec_fun(RecPun_Fun)
mtext("Rao diversity",adj=0,line=1)
mtext("Years since disturbance",side=1,adj=1,line=2)
mtext("Equivalent diversity",side=2,line=2.5)
```

Recruits functional diversity displayed a unimodal response to disturbance which maximum was positively correlated to disturbance intenity but that peaked at times unrelated to the disturbance treatment.

*Recruitment traits weighted means* 


Punctual and accumulated recruitment diversity of order 0, 1 and 2 were compared to a null recruitment model corresponding 

```{r Fig5, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Community weighted means (CWM) of the four disturbance treatment for the four leaf traits, the two stem traits  and the specific Hmax. Values reported correspond to the plot-level median obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes. Lines colors correspond to the disturbance intensity (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/Recruits_CWM")

par(mfrow=c(2,4),mar=c(3,3,2,1),oma=c(2,1,4,1),no.readonly=TRUE)
PlotCWM(Rec_CWM)
legendCWM()
mtext("Punctual recruits CWM",line=2,adj=0,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.6,cex=0.8,line=-0.2,outer=TRUE)
```


**Recruitment Turnover**

The abundance turnover of recruited trees compared to the initial stand inventoried before exploitation remained even for the 30 sampled years in the control plots \@ref(fig:Fig6). For the disturbed plot the abundance turnover displayed a unimodal response to disturbance, with maximum reached around 10 years after disturbance that was positively correlated to the disturbance intensity. The turnover had returned close to zero for all plots 30 years after disturbance.


```{r Fig6, out.width = '50%',echo=FALSE,fig.cap="Trajectories over the 30 sampled years of the abundance-based turnover between recruited trees and intial stands before disturbance. 50 repetitions of the taxonomic propagation process returned the distribution of trunover values, grey envelopes correspond to the 0.025 and 0.975 percentiles and lines to the median in green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/Turnover_toInit")

turnover(Turn)
par(mar=c(1,2,1,0))
mtext("Recruitment specific turnover compared to initial",side=3,adj=0,cex=1.2)
mtext("Years since disturbance",side=1,adj=1,line=-2)
mtext("Equivalent diversity",side=2,line=1)
```


# Discussion

From the 30 years of forest dynamics survey in the Paracou station we highlighted contrasting recruitment mechanisms between control and post-disturbance tropical forests. The difference among treatments were also significant which suggest an intensity threshold between the first and second disturbance treatment and could involve gaps size, the richness of exploited trees or the intervention density [@denslow1980].
> If we go for a model, maybe we can test the most relevant parameter!?

Disturbance increases the recruitment rate which here to increased the richness of recruited trees but impaired their evenness, all the more so that disturbance intensity is high. 
After disturbance the recruitment was then dominated by a restricted pool of species and Shannon and Simpson diversities decreased up to lower value than those of control plots.

For the 15 first years following disturbance the recruitement seemed driven by the growth of pre-disturbance saplings on hold of suitable growth condition that benefitted from the environmental changes. 
This translated by an increase in the functional diversity of recruits for all disturbance intensity and an even distribution of functional traits.

After this first recruitment phase, in the case of low disturbance intensity (T1 plots) the recruitment was highly nested within the community in place and displayed a richness and evenness similar to those of a random sampling null-model. 
This argues for the existence of some recruitment limitation that means the strive of a variety of species, even when not the most fitted for the environmental context, because the best competitors couldn't set in the disturbed area.
The key role of recruitment limitation in recovery dynamic was already highlighted for low disturbance intensity in tropical wet forests [@Hubbell1999;@Sheil2003;@Bongers2009] and supported the IDH as it mitigates competitive exclusion and maintains some inferior competitiors in the community.
Recruitment and dispersal limitation are consistent with the short dispersal distance observed for tropical trees, and it was already highlighted in Paracou through the genetic clumping observed for some pioneers [@Leclerc2015;@Scotti2015a].

In the case of high disturbance intensity, the species turnover was high and the functional diversity decreased rapidly, along a shift of the recruitment toward acquisitive functional strategies. 
This was particularly marked for functional traits of the leaf and stem and for life history traits (Leaf thickness, SLA, Wood Density and Hmax) which values represented a community of fast-growing species with a high resources acquisition strategy.
The time elpased between disturbance and the significance of those functional and diversity shifts suggests that the second recruitment phase reflects the germination of seeds from the soil seed bank at the moment of the disturbance. 
This recruitment would undergo a range of environmental filters strengthening the competitive exclusion and thus restricting the pool of species recruited [@Molino2001]. 

Eventually, we observed that disturbance induce particular recruitment mechanisms but it also set aside mechanisms at stake in natural plots.
Diversity of recruited community in undisturbed plots was higher than this observed for a random-sampling null model, suggesting mechanisms maintaining high diversity like negative density dependance [@Harms2000].


Previous studies already highlighted the significant diversity increase of communities as a whole (Mirabel et al., in.prep.) that here seems supported by a shift in recruitment dynamics in favor of acquisitive functional types [@TerSteege2001].

After 30 years stands functional characteristics had recover whenever the disturbance intensity but significant taxonomic diversity difference persisted when the disturbance intensity was high. 
Stands functioning would then recover faster than taxonomic composition, which questions the importance, the recovery and the role regarding diversity of functional redundancy in tropical forests.
In any case, the recruitment turnover compared to initial stand ends up close to zero.
This argues for the maintenance, or for the recovery in relation to the stand [@Anderson2007], of the before-disturbance environment.
It also supports the convergence of communities tree species taxonomic and functional composition in the long term [@Li2016].
Post-disturbance communities would then persist for more than 30 years as suggested by the remaining difference among treatments but the initial mechanisms seemed restored at this point, ensuring communities resilience.
On the same site, the different results of two studies 10 years [@Molino2001] and 20 years [@Baraloto2012a] already suggested the recovery toward initial taxonomic and functional condition of the stands in revealing a decreasing proportion of pioneers between the two periods.
This is however only valid for a single disturbance event, given that second recruitment phase seemed to rely on the soil seed bank a storage effect would modify the trajectories after another disturbance event.
We assume that in this case the competitive exclusion among dormant life-stage (seeds or even seedlings) would be harsher and bring more radical changes in the recuitment composition and functional profile.
Some functional shift toward more acquisitive, small seeded and globally disturbance resitant species building a more resilient community as it has been observed by [@Haddad2008] who linked the persistence of species to their ability to recover from disturbance which is the seed mass , growth rate linked to WD and SLA for plants.
Additionally to a somehow depleted seed bank this strongly questions the repeatability of observed trajectories after other disturbance events.

Our long-term study of diversity trajectories after disturbance desentangled the mechanisms underlying stands diversity. While in undisturbed forests mechanisms like negative density dependence enhance species diversity, any disturbance induce new mechanisms depending on the disturbance intensity.
In any case, communities short term response first reflects the growth of a variety of already grown sapling which increases the taxonomic and functional evenness of the community.
When the disturbance is low, recruitment limitation  is balanced against environmental filters and induce a recruited community mirroring the stand in place while selecting more acquisitive functional type.
When the disturbance is intense however the species turnover is high and induce a recruited community significantly dominated by acquisitive functional types and a restricted pool of species.
In both cases plots recruitment is specific, because it is determined by the surrounding stand in case of low disturbance or because the competitive exclusion randomly selects a restricted pool of species after intense disturbance.
At the landscape scale then disturbance increases the diversity through the "between patch" difference [@denslow1987;@Chesson2000; @Sheil2003].

> Connell 1978, refute the "equal chance HP because trees don't recruit equally"
> Contrary to Hubbell and Molino: the question is "do gaps allow specific species to establish which would thrive otherwise?" [@Sheil2003]. The turnover answers that yes it does
<!-- Baraloto, 2012: on juveniles the differences were still at stake. --> 

----

`r if (knitr:::is_html_output()) '# References {-}'`
