---
title: "30 Years of Recruitment in Tropical Forest After Selective Logging"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric MARCON
    authsuperscript: 1
  - name: Bruno HERAULT
    authsuperscript: 2
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB (Institut National Ploytechnique Félix Houphoüet Boigny) 
    address: Yamoussoukro, Ivory Coast
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >
  Résumé de l'article.
keywords: [mot-clés, séparés par des virgules]
journalinfo: Hal 00679993
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
bibliography: references.bib
lang: english
keywordlabel: Mots-clés # Keywords
corrauthorlabel: Contact # Corresponding author
output: 
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::gitbook: default
  bookdown::word_document2: default
---

<!-- Options de knitr et de R (librairies, ...) --> 

```{r Options, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(cache=F, warning=FALSE, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=40),fig.align="center")
# Installation des packages si nécessaire et chargement
Library <- function(Packages) {
  InstallAndLoad <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {install.packages(Package, repos="https://cran.rstudio.com/")}
    require(Package, character.only = TRUE)
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Ajouter les packages nécessaires ici
Library("kableExtra")
#source("Scripts/GithubPages.R")
source("Scripts/DrawingTools.R")
```

# Introduction

Determining the response of tropical forests to disturbance is a key to predict their fate in a global change context.
A vast literature has successfully modeled the response of tropical forest dynamics, carbon stocks and fluxes to anthropogenic and natural disturbances [@Gourlet-Fleury2000;@Putz2012;@Martin2015;@Piponiot2016]. 
However, regarding diversity, similar attempts have been hindered by both the huge biological diversity and the scarcity of long-term monitoring. 
Although the community composition response to disturbance was identified for common species assemblages, this approach is usually confined to few commercial and valuable species [@Sebbenn2008;@Rozendaal2010;@Vinson2015]. 
Given the complexity of adressing simultaneously the response of dozens of species, community-scale approaches have been developped [@DeAvila2016].

The post-disturbance community dynamic is largely determined by the recruitment process, i.e. from seed production, dispersion and germination to seedlings' and saplings' growth until a threshold, usually 10cm DBH in tropical forest plots.
Major ecological processes, such as environmental filtering or limiting similarity, occur before formal recruitment and determine the taxonomic composition of trees eventually integrating the forest community [@Ackerly2003]. 
The recruitment process is inherently linked to the disturbance regime that locally changes ecosystem's biotic and abiotic conditions and creates opportunities for pioneer species to strive in the long term rather than undergoing the competitive exclusion for resources in mature patches [@Denslow1980]. 
This relies on the Intermediate Disturbance Hypothesis (IDH) that explains the maintenance of tropical forests biodiversity by the patchy variability of environmental conditions in space and time [@Guitet2018]. 
This disturbance-induced environmental variability, when not too intense, is assumed to enlarge the ecological range of species present in the community [@Molino2001;@Bongers2009] and shapes its taxonomic diversity, vegetative structure, physiology as well as carbon, nutrients, and water cycles [@Anderson-Teixeira2013]. 
Empirical tests of the IDH in tropical rainforests, though, proved hard to succeed and yielded controversial results [@Hubbell1999;@Molino2001;@Sheil2003].

During post-disturbance times, the shift from resource-acquisitive to resource-conservative ecological strategies may be detected in leaves (leaf thickness, toughness, chlorophyll content and specific area) and stem (wood specific gravity and bark thickness) and life-history traits (maximum height at adult stage and class of seed mass) [@Wright2004;@Chave2009b;@Herault2011]. 
The relative importance of recruitment of new individuals and of mortality of distrubance survivors will shape the new forest and its functioning. 
Given that disturbance survivors largely mirror the pre-disturbance forest composition [@Piponiot2018], predicting the recruitment composition and diversity trajectories would be a major step towards the prediction of the future of tropical forest in a changing global environment where disturbance are expected to become more and more frequent. 
This would give insights into the resilience of this hyperdiverse ecosystems, elucidate the determinism, or not, of tropical forests trajectories, test the convergence after disturbance of taxonomic and functional communities towards initial state and also help future adaptative conservation strategies [@Diaz2005;@Gardner2007;@Schwartz2017].

In this paper we follow the fate of a recruited tree communities (60121 individuals) over 30 years on a disturbance gradient, with 10 to 60% of forest biomass removed. 
We assessed the taxonomic as well as functional diversity of recruited trees, using a large functional trait database covering, the leaf, wood and life-history spectra. 
We aimed to (i) assess the role of environmental filtering selecting the recruited trees according to their competitivity for resource acquisition, (ii) resolve the convergence of communities and the maintenance of taxonomic composition in the long term, and (iii) determine the global resilience of the ecosystem.

# Material and Methods


## Study Site

The Paracou station is located in a lowland tropical rain forest in French Guiana (5°18'N and 52°53'W). Climate is tropical wet with mean annual precipitation averaging 2980 mm.y^-1^ (30-y period) and a 3-months dry season (< 100 mm.months^-1^) from mid-August to mid-November, and a one-month dry season in March [@Wagner2011]. Elevation ranges from 5 to 50 m and mean annual temperature is 26°C. Soils are thin acrisols over a layer of transformed saprolite with low permeability generating lateral drainage during heavy rains. The disturbance experiment spread over a network of twelve 6.25ha plots (Table \@ref(tab:Tab1)) that underwent three disturbance treatments in 1986-1987 [@Herault2018]. Dominant families are Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae. 

```{r Tab1, echo=FALSE}
Intervention <- data.frame(
  Treatment=c("Control", "T1", "T2", "T3"),
  Timber=c("", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha",
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha"),
  Thinning=c("", 
             "", 
             "DBH $\\geq$ 40 cm, non-valuable species, $\\approx$ 30 trees/ha", 
             "DBH $\\geq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha  "),
  Fuelwood=c("", 
             "", 
             "", 
             "40 cm $\\leq$ DBH $\\leq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha"),
  AGBLost=c(0, "$[12\\%-33\\%]$", "$[33\\%-56\\%]$", "$[35\\%-56\\%]$")
)
names(Intervention)[5] <- "\\%AGB lost"
knitr::kable(Intervention, caption="Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.", longtable = FALSE, booktabs = TRUE, escape = FALSE,format="latex") %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = TRUE)
```

## Inventories Protocol and Dataset Collection 
All trees above 10 cm DBH were mapped and measured annually since 1984. During inventories, trees were first identified with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. Botanical campaigns have been carried out every 5 to 6 years from 2003 onwards. This raised methodological issues as vernacular names usually correspond to different botanical species, resulting in significant taxonomic uncertainties that were propagated to composition and diversity metrics. Vernacular names were replaced through multinomial trials $M_v\Big(\big[s_1, s_2, …, s_N\big],\big[\alpha_1, \alpha_2,…, \alpha_3\big]\Big)$  based on the observed association probability $\big[\alpha_1, \alpha_2,…, \alpha_3\big]$ between each vernacular name *v* and the species $\big[s_1, s_2, …, s_N\big]$ recorded in the inventory. See appendix 1 and @Aubry-Kientz2013 for the detailed methodology. To avoid remaining identification caveats, the simulated botanical inventories were reported at genus level.

Six functional traits, representing leaf economics (leaves thickness, toughness, total chlorophyll content and specific leaf area, the leaf area per unit dry mass), wood economics (wood specific gravity and bark thickness) and life history traits (maximum specific height and seed mass), come from the BRIDGE project [^1] where trait values were measured on nine french guianan forest plots, including two in Paracou. Missing trait values (X%) were filled using multivariate imputation by chained equation (mice) restricted to samples pertaining to the next higher taxonomic level, in order to account for the phylogenetic signal of the functional traits (refs MICE). As seed mass information corresponds to a classification into mass classes, no data filling process was applied so analysis were performed only considering the 414 botanical species of the seed mass dataset.

[^1]: http://www.ecofog.gf/Bridge/

Functional trajectories were estimated with the Rao quadratic entropy using community weighted means (CWM) [@Diaz2007;@Garnier2004]. Seed mass trajectories were reported by the proportion of each class recorded in the inventories.
All composition and diversity metrics are the average obtained after 50 iterations of uncertainty propagation (taxonomy and trait values). 


## Recruitment trajectories

We split the forest community in 'survivors, i.e. trees that survived the disturbance,  and post-disturbance recruited trees. 
Two recruitment metrics were examined: on the one hand the "punctual recruitment" by 2-year intervals after disturbance, on the other hand all recruited trees since disturbance, hereafter “accumulated recruits”. 
The taxonomic diversity was assessed through Richness and the Hill number translation of Shannon and Simpson indices [@Hill1973;@chao2015estimating;@Marcon2015b]. 
These three indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (*q*), which grasps the balance between richness and evenness in the community, with common species weighting more than rare ones when *q* increases. 
The similarity between the recruited trees and the pre-disturbance forest was measured with the turnover metrics detailed in @Podani2013a. 
To determine whether recruitment trajectories ensued from a pure random process, observed trajectories were compared to those generated by 50 repetitions of a random null model shuffling individuals among plots while preserving species abundance and plots’ tree density.

To draw plots trajectories we applied a moving average with a one step window allowing to mitigate the heterogeneity of inventory protocols between years.

# Results

## Recruitment Diversity

### Taxonomic diversity

Punctual recruits' diversity followed a consistent trajectory among disturbance treatments with first higher richness and lower evenness than in control plots and then equivalent richness and lower evenness (Figure (\@ref(fig:Fig1)). 
The richness of accumulated recruits was higher for the disturbed plots, consistently with the increase of recruited trees following disturbance, but their evenness was equivalent or lower, specifically for the most disturbed plots.

```{r Fig1, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Trajectories of Richness, Shannon and Simpson diversity for 2-years laps punctual  recruitment (upper panels) and divergence to null model (lower panels). Lines colors refer to the perturbation regime: green for control, blue for T1, orange for T2 and red for T3 disturbance treatments. Plain lines correspond to the median observed after uncertainty propagation and are given along with the 95\\% confidence interval (grey envelope).", fig.env="figure*"}
load("DB/RecruitmentPunctual");load("DB/RecruitmentPunctual_Nullmodel_Diff")

TrajectoryDiffNull(RecDB=RecPun,RecDB_Diff=RecPun_DiffNull)

```

Punctual and accumulated recruitment diversity of orders 0, 1 and 2 were then compared to a null recruitment model with random sampling. 
In control plots, punctual recruitment values were constantly equivalent or higher than with the null model (Figure \@ref(fig:Fig1)). 
In disturbed plots, punctual recruitment richness and evenness remained lower than of the null model whenever the disturbance treatment.
Accumulated recruitment richness (order 0) and evenness (order 2) were higher or equivalent to those of the null model for plots T1 and some plots T2 but lower for plots T3 and a plot T2. 

### Functional diversity

```{r Fig2, fig.asp=1.1/1, echo=FALSE,fig.cap="Functional diversity of punctual recruited trees from the considered functional traits and divergence to null model. Values reported correspond to the plot-level median and the 95\\% confidence interval obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes and 50 run of the null model. Lines colors correspond to the logging treatment initially applied (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/RecruitmentPunctual_FunctionalN");load("DB/RecruitmentPunctual_Functional_Nullmodel_Diff")

TrajectoryRec_fun(RecDB_fun=RecPun_Fun,RecDB_fun_Diff=RecPun_Fun_DiffNull)
```

Recruits functional diversity lower after disturbance for 15 to 20 years period before increasing to a similar or higher value as observed for control plots (Figure \@ref(fig:Fig3)). 
Trajectories of recruited trees in the functional spaces showed the dominance after disturbance of species displaying large exchange surface area and light tissues (high SLA, low leaf toughness and low wood specific gravity) (Figure \@ref(fig:Fig3)).
All traits trajectories displayed univariate CWM trajectories, except SLA and leaf thickness that displayed a unimodal trajectory. 

```{r Fig3, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Community weighted means (CWM) of the four disturbance treatment for the four leaf traits, the two stem traits  and the specific Hmax. Values reported correspond to the plot-level median obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes. Lines colors correspond to the disturbance intensity (green for control, blue for T1,orange for T2 and red for T3).", fig.env="figure*"}
load("DB/Recruits_CWM")

par(mfrow=c(2,4),mar=c(3,3,2,1),oma=c(2,1,4,1),no.readonly=TRUE)
PlotCWM(Rec_CWM)
legendCWM()
mtext("Punctual recruits CWM",line=2,adj=0,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.6,cex=0.8,line=-0.2,outer=TRUE)
```


## Recruitment Turnover

In control plots, species turnover remained highly stable for the 30 sampled years (Figure \@ref(fig:Fig4)). 
In disturbed plots, turnover displayed a unimodal response to disturbance, with maximum reached around 15 years and with a value positively correlated to the disturbance intensity ($\rho_{spearman}=0.91$). The turnover trajectory returned close to zero for all plots 30 years after disturbance.


```{r Fig4, echo=FALSE,fig.cap="Trajectories over the 30 sampled years of the abundance-based turnover between recruited trees and intial communities before disturbance. Grey envelopes correspond to the 0.025 and 0.975 percentiles of the uncertainty propagation procedue and lines to the median in green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/Turnover_toInit")

turnover(Turn)
par(mar=c(1,2,1,0))
mtext("Recruitment specific turnover compared to initial",side=3,adj=0,cex=1.2)
mtext("Years since disturbance",side=1,adj=1,line=-2)
mtext("Equivalent diversity",side=2,line=1)
```


# Discussion

From the 30 years of forest dynamics survey in the Paracou station, we highlighted contrasting recruitment patterns depending on disturbance intensity. 
Disturbance increased the recruitment rate which subsequently increased the richness of recruitment but impaired its evenness, all the more so that disturbance intensity was high. After disturbance the recruitment was dominated by a restricted pool of species and Shannon and Simpson diversities decreased down to lower values than those of control plots.

## On the underlyings of the hump-shaped trajectories

The punctual recruitment richness, some key functional traits (SLA and bark thickness) and the species turn-over trajectories exhibited hump-shaped curves so that, for the 10-15 first years following disturbance, recruitment processes seemed driven by the growth of pre-disturbance saplings benefitting from the environmental changes and alleviated competition [@Herault2010].
After this recruitment progressively incorporated true recruits, *i.e.* individual trees that had germinated after the disturbance, undergoing competition again and environmental selection. 
This translated first by a stable functional diversity of the recruitment, equivalent as this of control plots, and low difference to random null model, at least for the lower intensities. 
According to the SLA trajectory (Figure \@ref(fig:Fig4)), these true recruits dominating the recruitment 12 years after disturbance are mainly pionneer species.

After this first recruitment phase, different trajectories were observed according to the disturbance intensity.
In the case of low disturbance intensity (T1 plots) the recruitment resembled pre-disturbance species composition.
Besides it displayed on the one hand decreasing eveness compared to control plots and random models, translating selective recruitment, and on the other hand increasing functional diversity reaching the levels of control plots or above, translating some competitive exclusion. 
The second recruitment phase at this level of disturbance is then dominated by pioneers and light demanding species either competing with each other and resulting in a functionally diversified community, or preserved from the overwhelming settlement of hard pioneers by recruitment limitation [@Hubbell1999;@Sheil2003;@Bongers2009].
Recruitment limitation of hard pioneers, mitigating competitive exclusion and maintaining some inferior competitiors in the community, would be consistent with the short dispersal distance observed for tropical trees, specifically in Paracou with the genetic clumping of some pioneers [@Leclerc2015;@Scotti2015a].
In the case of high disturbance intensity, the species turnover remains very high after the initial 10 year phasis and the functional diversity decreased sharply towards a restricted range of acquisitive functional strategies.  
Both recruitment phase were marked for nearly all functional traits of the leaf and stem and for life history traits, reflecting the dominance of high resources acquisition strategies [@Wright2004;@Chave2009b;@Herault2011;@Reich2014].
However the first recruitment phase corresponded to a sharp increase in SLA and bark thickness, reflecting the dominance of short-lived pionneers, while the second recruitment phase corresponded to a decrease in leaf thoughness, Wood Specific Gravity and Hmax relecting the incorporation of long-lived pioneers.

## On the resilience of the recruitment process

Thirty years after disturbance, recruitment functional diversity and richness had recovered equivalent levels as those observed for the plots short after after disturbance and equivalent to those of control plots.
The functional and taxonomic composition of the recruitment and the evenness of recruited species distribution, though, remained lower for all disturbed plots, along with and weighted means of all functional traits was remained altered thirty years after disturbance. 
Would this mean that ecosystem functioning recover faster than taxonomic composition? 
The link between functional traits and ecosystem functionning is not trivial but our results may highlight the key role of functional redundancy in ecological systems and also suggest.
>> that the recovery trajectory may be, more than commonly thought, upon dependence of the pre-disturbance functional ecosystem signature [@Herault2018]. -> What makes us say that?

Whatever the disturbance intensity, the recruitment turnover compared to initial stand ended up close to zero. 
This argued for the high dependence of the composition recovery trajectory on the pre-disturbance ecosystem characteristics [@Anderson2007]. 
In other words, initial compositional variation caused tree communities to remain divergent in taxonomic composition, even though these same tree communities strongly converged in functional space [@Fukami2005]. 
This makes these commmunities both functionally and taxonomically resilient despite the settlment of long lived pioneers which make it a long term process for high disturbance intensity.
>> Not that much because the long lived pioneers probably remain long after disturbance, it would be in the long term.

Our results extend previous ones from the Paracou experiment, 10 years [@Molino2001] and 20 years [@Baraloto2012a] after disturbance that suggested the recovery towards pre-disturbance taxonomic and functional composition. 
This is however only valid for a single disturbance event, given that second recruitment phase assumed to rely on the seed bank triggers a storage effect likely to modify the recruitment trajectories in case of a new disturbance event. 
In this hypothetical case, the competitive exclusion among dormant life-stage (seeds or even seedlings) would be harsher and would bring more radical changes in the recruitment composition and functional profile.

## Conclusion

Our long-term study of diversity trajectories after disturbance disentangled the mechanisms underlying the forest trajectory in the diversity space. 
While, in undisturbed forests, mechanisms like negative density dependence enhanced species diversity, disturbance induced new mechanisms depending on its intensity. 
In the short-term communities response was always driven by the enhanced growth of grown saplings benefitting from alleviated competition and low selective pressure resulting in increased communities taxonomic and functional evenness. 
In the long-term communities trajectory strongly depended on the disturbance intensity. 
At low intensity competition resumed to shape a functionally diversified and even community preserved from the over-whelming hard pioneers by recruitment limitation.
For high intensity, strong environmental filters maintained high species turnover and a recruitment restricted to long lived pionneers. 
Communities composition and functioning proved consistently resilient to single disturbance events but the functional redundancy and the diversified seed bank of mature forests proved altered and consequently communities resilience themselves.

----

`r if (knitr:::is_html_output()) '# References {-}'`
