---
title: 30 Years of Recruitment in Tropical Forest After Selective Logging 
author:
  - name: Ariane Mirabel
    authsuperscript: 1*
  - name: Eric Marcon
    authsuperscript: 1
  - name: Bruno Hérault
    authsuperscript: 2
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: to be completed
    address: >
      Street address,
      Zip code,
      Country.
corrauthor:
    email: prenom.nom@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >
  Résumé de l'article.
keywords: [mot-clés, séparés par des virgules]
journalinfo: Hal 00679993
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{amsmath}
bibliography: references.bib
lang: french # english
keywordlabel: Mots-clés # Keywords
corrauthorlabel: Contact # Corresponding author
output:
  bookdown::gitbook
#   bookdown::pdf_book:
#     base_format: EcoFoG::article
#     keep_tex: TRUE
#  bookdown::word_document2
---

<!-- Options de knitr et de R (librairies, ...) --> 

```{r Options, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(cache=TRUE, warning=FALSE, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=40),fig.align="center")
# Installation des packages si nécessaire et chargement
Library <- function(Packages) {
  InstallAndLoad <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {install.packages(Package)}
    require(Package, character.only = TRUE)
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Ajouter les packages nécessaires ici
#source("Scripts/GithubPages.R")
source("Scripts/DrawingTools.R")
```

<!-- Début du document ci-dessous --> 


# Introduction

Decades of studies retained demographic processes driving trees recruitment as the main determinant of forest dynamics. 
As stated by the Intermediate Disturbance Hypothesis (IDH), a high biodiversity in communities would be maintained by the variability of environmental conditions which would favor and enhance the growth of a variety of species along time which makes the recruited trees to be very diverse. 
Similarly, the high diversity of tropical forests is assumed to be maintained by the tree fall gaps dynamics which impulses a different recruitment regime between enlightened and dyer gaps and shaded, wet undercover areas. 
Recruitment dynamics are thus crucial to understand, as determinant of the forest-to-be and driver of community assemblage.
A large literature investigated forests population dynamics and successfully modeled forest response to disturbance. 
Based on stands structure observed in the field, these models proved relevant predictors of carbon sequestration, commercial stocks and diameter distribution dynamics in the short and in the long term \ref{@Gourlet-Fleury2005}. 
Predicting in this way species population recovery and subsequently stands diversity after disturbance however seems tricky as the growth and survival dynamics in the early life stage, far from being well understood, are highly variable among species which makes the recruitment of a tree too infrequent an event and really poorly correlated to the presence of similar species at the time considered. 
Attempts were successfully conducted to model the regeneration of single species however the response of the whole recruited community after disturbance could not reasonably be assessed via similar methods which would require the combination of accurate knowledge of each species ecology and behavior.

-	Post logging dynamics have been identified but the underlying mechanisms remain unclear, and it is interesting because it gives clues on the functioning of forests (which filters are at stake and how they apply on the ecosystems) 
-	We were interesting in the successional rate along which recruitment functional diversity varies: do we already see the new generation of functional type selected, compared to mature (old growth) forest

-	Do the stand return to initial condition, is the cycle achieved?

-	Hypotheses: 

  1) The more intense the logging, the most significant the impact on composition and diversity metrics. 

  2) We expect an increase in the stand diversity (Intermediate disturbance hypothesis) or no change in richness, at least for low logging intensity

  3) Exploitation would change the functional composition toward more light-demanding strategies 

  4) We expect the diversity to follow a cyclic dynamic, as observed for stands biomass, and the ecosystem functioning to be resilient, at least for low intensity treatment


# Material and Methods

*Study Site* 
Analyses were based on the inventories conducted at the Paracou station in French Guiana (5°18'N and 52°53'W) in a lowland tropical rain forest. 
The site corresponds to a tropical wet climate with mean annual precipitation averaging 2980 mm.y-1 (30-y period) with a 3-months dry season (< 100 mm.months-1) from mid-August to mid-November, and a one-month dry season in march. 
Elevation ranges between 5 and 50 m and mean annual temperature is 26°C. 
Soils correspond to thin acrisols over a layer of transformed saprolite with low permeability generating lateral drainage during heavy rains.
The experiment corresponds to a network of twelve 6.25ha plots where three logging treatments representing a gradient of logging intensity were applied from 1985 to 1987. 
Logging treatments were attributed according to a randomized plot design with three replicate blocks of four plots. 
The logging treatments correspond to averages of 10 trees remove per hectare with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees above 40 cm DBH for treatment 3 (T3). 
Treatments T2 and T3 included besides logging of 10trees/ha the thinning by poison girdling.

*Inventories Protocol and Dataset Collection* 
The study site corresponds to a tropical rain forest with a dominance of Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae. 
In the twelve experimental plots all trees above 10 cm DBH are mapped and measured annually since 1984, which allows to survey the trees recruited annually. 
During the inventories trees are first identified with a vernacular name, assigned by the field team, and then with a scientific name, assigned by a botanist during regular botanical campaigns. 
In 1984, specific vernacular names were given to 62 commercial or common species whereas two identifiers separating palm trees and trees were given to all other species. 
Botanical campaigns to identify all trees at the species level started in 2003 and continued every 5 to 6. 
In total from 1984 to 2016 around 850 species were identified in the 12 plots. 
For the trees that were identified at the species level vernacular names attributed in the first place generate some taxonomic uncertainty as they may correspond to a variety of botanical species. 
To account for this taxonomic uncertainty in the composition and diversity metrics we used a Bayesian framework estimating the expectancy and variance of composition and diversity metrics. 
The frameworks is based on a set of complete inventories extrapolated from the initial one through the replacement of vernacular name by botanical species. 
These replacement were made through a multinomial distribution M~v~([s~1~, s~2~, …, s~N~] ,[α~1~, α~2~,…, α~N~]) based on the observed association probability [α~1~, α~2~,…, α~N~] between each vernacular name v and the species recorded in the inventory [s~1~, s~2~, …, s~N~]. 
See appendix 1 for the detailed methodology. 
Diversity metrics correspond to the expectancy obtained after 50 iteration of the taxonomic uncertainty propagation framework. 

*Recruitment Survey* 
To assess the demographic dynamics along the 30 sampled year we divided the inventoried trees between trees recorded before the logging in 1984 and the trees recruited afterward. 
Analysis first surveyed the communities of trees recruited every 5 years-laps after exploitation, hereafter called “punctual recruitment”. 
Then to assess the diversity of the stands under settlement analysis surveyed the communities of recruited trees accumulated since exploitation, hereafter called “accumulated recruits”.
To tackle the unequal number of recruited trees between treatments, which would mislead the diversity with the effect of sampling size, the diversity was assessed as the average measure obtained from the resampling of 50 trees in the 5 years – laps recruited communities. 
The diversity of accumulated recruits corresponded to the accumulation of resampled trees.
Taxonomic diversity was assessed through Richness and the Hill number translation of Shannon and Simpson indices. 
These three indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (q). 
This order q grasps the balance between richness and evenness in the community as it determines the emphasis on common species in the diversity metric, with common species weighting more than rare ones when q increases. 
The similarity between the recruited trees and the old growth forest was measured with the turnover metrics detailed in @Podani2013a. 
The metric used correspond to the relativized abundance replacement, the sum of abundance in one site that is replaced by completely different species, normalized by the maximum abundance shared by the two communities.
T_ab=(∑_(i=1)^n▒|x_ia 〖-x〗_ib | -|∑_(i=1)^n▒〖x_ia-∑_(i=1)^n▒x_ib 〗|)/(∑_(i=1)^n▒max{x_ia;x_ib } )

To determine whether trees recruitment ensued from a random process the observed diversity trajectories were compared to those generated by 50 repetitions of stochastic null models.
The null model for taxonomic diversity randomly shuffled individuals among plots while preserving species abundance and plots’ tree density. 


# Results

Over time, 104423 trees and 697 botanical species were recorded. 

**Recruitment Diversity** 


*Punctual and accumulated recruits’ diversity*
The diversity of trees recruited in T1, T2 and T3 logged plots constituted a community with lower richness and lower evenness than observed for the control plots. 
The OLS analysis proved a significant time-dependence of the punctual recruitment diversity for the three diversity order, Richness, Shannon and Simpson and a significant effect of the logging treatment (P<0.01 in any case). 
It besides proved a significant time-dependence of the accumulated recruitment diversity for the three order and a significant effect of the logging.


```{r Fig1, out.width = '80%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Trajectories for the 30 sampled years of Richness, Shannon and Simpson diversity for the **(a)** 5-years laps punctual recruitment and **(b)** accumulated recruits. The diversity measured correspond to the average diversity of recruited trees within 5 years – laps. Line colors represent the logging treatment (green for control, blue for T1,orange for T2 and red for T3). The 0.025 and 0.975 percentile correspond to the variance observed for 50 iteration of the taxonomic uncertainty propagation."}
load("DB/RecruitmentAccum");load("DB/RecruitmentPunctual")

par(mfrow=c(2,3),mar=c(2,3,1,1),oma=c(2,3,1,1),no.readonly=TRUE)
TrajectoryRec(RecPun,0.8)
TrajectoryRec(RecAccum,1)
mtext("(a) Richness",at=0.1,line=-0.6,outer=TRUE);mtext("(b) Shannon",at=0.44,line=-0.6,outer=TRUE)
mtext("(c) Simpson",at=0.76,line=-0.6,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.85,cex=0.8,line=1,outer=TRUE)
mtext("Equivalent diversity",side=2,padj=1,cex=0.8,line=0.8,outer=TRUE)
mtext("Punctual",side=2,adj=0.8,line=1.75,outer=TRUE)
mtext("Accumulated",side=2,adj=0.2,line=1.75,outer=TRUE)

```


*Comparison to null model*
Punctual and accumulated recruitment diversity of order 0, 1 and 2 were compared to a null recruitment model. 
The null model randomly sampled the adequate number of recruited species, as observed in the real 5-years laps inventory. 

```{r Fig2, out.width = '100%', fig.asp = 1.5/2, echo=FALSE, fig.cap="Punctual recruitment diversity trajectories for the Richness, Shannon and Simpson diversity compared to the null model trajectories. Values reported correspond to 0.5 quantile of plot-level median diversity obtained after 50 repetition of the taxonomic uncertainty propagation process. Lines colors correspond to the logging treatment initially applied (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/RecruitmentPunctual_Nullmodel")

TrajectoryRec_vsNull(RecPun,RecPun_Null)
mtext("Punctual recruitment",line=1.2,adj=0,outer=TRUE,cex=1.2)
mtext("Richness",at=0.1,line=-0.5,outer=TRUE);mtext("Shannon",at=0.44,line=-0.5,outer=TRUE)
mtext("Simpson",at=0.76,line=-0.5,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.85,cex=0.8,line=1,outer=TRUE)
mtext("Equivalent diversity",side=2,padj=1,cex=0.8,line=1.5,outer=TRUE)

```

```{r Fig3, out.width = '100%', fig.asp = 1.5/2, echo=FALSE, fig.cap="Accumulated recruitment diversity trajectories for the Richness, Shannon and Simpson diversity compared to the null model trajectories. Values reported correspond to 0.5 quantile of plot-level median diversity obtained after 50 repetition of the taxonomic uncertainty propagation process. Lines colors correspond to the logging treatment initially applied (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/RecruitmentAccum_Nullmodel")

TrajectoryRec_vsNull(RecAccum,RecAccum_Null)
mtext("Accumulated recruitment",line=1.2,adj=0,outer=TRUE,cex=1.2)
mtext("Richness",at=0.1,line=-0.5,outer=TRUE);mtext("Shannon",at=0.44,line=-0.5,outer=TRUE)
mtext("Simpson",at=0.76,line=-0.5,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.85,cex=0.8,line=1,outer=TRUE)
mtext("Equivalent diversity",side=2,padj=1,cex=0.8,line=1.5,outer=TRUE)

```


```{r Fig4, out.width = '100%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Community weighted means (CWM) of the four disturbance treatment for the six leaf and stem functional traits considered. Values reported correspond to 0.5 quantile of plot-level median obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes. Lines colors correspond to the logging treatment initially applied (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/Recruitment_CWM")

PlotCWM(Rec_CWM)

```

```{r Fig5, out.width = '100%', fig.asp = 1/3, echo=FALSE,fig.cap="Functional richness, evenness and dispersion from the considered functional traits. Values reported correspond to 0.5 quantile of plot-level median obtained after 50 repetition of the taxonomic uncertainty propagation and the functional database gap-filling processes. Lines colors correspond to the logging treatment initially applied (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/Punctual_Fevenness");load("DB/Punctual_Frichness")

par(mfrow=c(1,3),mar=c(2,2,1,1),oma=c(2,2,1,1),no.readonly=TRUE)
FDiversity(Punctual_FRich)
FDiversity(Punctual_FEve)
mtext("(a) Richness",at=0.1,line=-0.6,outer=TRUE);mtext("(b) Evenness",at=0.44,line=-0.6,outer=TRUE)
mtext("(c) Dispersion",at=0.76,line=-0.6,outer=TRUE)
mtext("Years since disturbance",side=1,at=0.85,cex=0.8,line=1,outer=TRUE)
mtext("Equivalent diversity",side=2,padj=1,cex=0.8,line=1.5,outer=TRUE)

```


**Recruitment Turnover**

The abundance turnover of recruited trees compared to the initial stand inventoried before exploitation remained even for the 30 sampled years in the control plots whereas it showed a first significant increase for 15 years after exploitation in all logged plots. 
The increase depended on the logging intensity with higher abundance turnover after the most intense exploitation. 
The turnover then decreased in all plots from 15 years after exploitation and plots of all treatments displayed a turnover equal to the control plots from 25 years after exploitation.


```{r Fig6, out.width = '60%',echo=FALSE,fig.cap="Trajectories over the 30 sampled years of the abundance-based turnover between recruited trees and intial stands before disturbance. 50 repetitions of the taxonomic propagation process returned the distribution of trunover values, grey envelopes correspond to the 0.025 and 0.975 percentiles and lines to the median in green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/Turnover_toInit")

turnover(Turn)
par(mar=c(1,3,2,0))
mtext("Recruitment specific turnover compared to initial",side=3,adj=0,line=1,cex=1.2)
mtext("Years since disturbance",side=1,adj=1,line=-2)
mtext("Equivalent diversity",side=2,line=1.5)

```


# Discussion

-	The pool of recruited trees is restricted after exploitation, suggesting the involvement of additional filters (biotic or abiotic) compared to natural forests. 

-	This is confirmed with the comparison to null model: the recruitment after exploitation is more diverse than in natural forests.

-	The punctual diversity is lower after exploitation but the diversity of the accumulated recruitment is higher: the filters change along time.

-	The IDH all the more strong the filters are different compared to natural forests so the diversity increases with the mix of after logging and old-growth species + with the changes of after logging species. 
It is confirmed with functional analysis.

-	The change in filters is actual until 20 or 25 years after logging: from then on the species recruited are nested within the initial community.

-	The stand is on the way to recover its initial condition, whenever the treatment, but this will probably be much longer than the 30 years envisaged for the logging rotations. 
Baraloto, 2012: on juveniles the differences were still at stake.

-	The big question is then: would we observe the same dynamic after a second cycle or is the seed bank depleted after the recruitment of old-growth species. 
A second rotation would be lethal for the stand composition recovery.


`r if (knitr:::is_html_output()) '# References {-}'`
